The capsule and genetic background, rather than specific loci, strongly influence pneumococcal growth kinetics

Version 1

Abstract

Abstract

Bacterial growth characteristics intrinsic to each strain can impact and influence gene expression, antibiotic susceptibility, and disease pathogenesis. However, little is known about specific genomic variations that influence these bacterial growth features. Here, we investigate the impact of Streptococcus pneumoniae genetics on its in vitro growth features to shed light on genes that may be important targets in the development of vaccines and therapeutics. We use statistical models to estimate growth features and show that they varied significantly across capsular serotypes and lineages, strongly correlated with the phylogeny, and were highly heritable, highlighting a strong genetic basis. Despite this, genome-wide association studies revealed no specific genomic loci statistically associated with the growth features independently of the genetic background, including those in the capsular polysaccharide synthesis locus. Our findings show that the serotype and lineage, and possibly a combination of genomic loci, influence intrinsic pneumococcal growth kinetics, which may have implications on pneumococcal disease pathogenesis.

Introduction

Once termed “the captain of the men of death” by Sir William Osler in the 20th century, Streptococcus pneumoniae, or the pneumococcus, remains a significant cause of life-threatening invasive diseases, such as sepsis, pneumonia, and meningitis1, despite the widespread implementation of life-saving serotype-specific polysaccharide pneumococcal conjugate vaccines (PCVs)2. The high capsular diversity, amounting to ∼100 serotypes with varying potential for colonising the nasopharynx and causing invasive diseases3,4 and mortality5–7, demonstrates the remarkable genetic diversity of the pneumococcus. Such high genetic diversity reflects the plasticity of the pneumococcal genome, which readily swaps genetic material with other strains through recombination and horizontal gene transfer8. However, the impact of pneumococcal genomic variation on its phenotypes, especially those associated with disease pathogenesis and clinical manifestations, remains less understood.

The pneumococcus possesses an arsenal of virulence factors. The extracellular polysaccharide capsule is the main virulence factor critical for transmissibility, survival, and pathogenicity in the human host9–11. Recent studies in the pneumococcus and other bacterial pathogens suggest that variability in growth kinetics impacts disease pathogenesis, clinical manifestation, and antibiotic therapy. For example, differential pneumococcal growth features appear to correlate with fitness for colonisation and invasive disease,12 clinical manifestation12–14 and response to environmental conditions14–16. Studies of other bacterial pathogens suggest that growth rate influences gene and protein expression17,18, adhesion19, bacterial competition20, plasmid replication21, bacteriophage dynamics22, and antibiotic accumulation and efficacy23–25. Similarly, experimental and mathematical modelling studies have demonstrated that bacterial growth correlates with bacterial load and host-mediated cell lysis and impacts dispersal to different host cells and tissues during systemic infection26. Previous studies of pneumococcus and other bacteria have suggested the influence of environmental factors15,16 and bacterial-specific factors, including capsular serotype14,27 and the presence of plasmids28 on bacterial growth phenotypes. These growth phenotypes may consequently impact invasive pneumococcal disease pathogenesis12–14. However, the specific pneumococcal genomic variations affecting growth kinetics remain less understood.

Studying bacterial growth kinetics sheds light on the influence of genes, genomic variation, and growth media conditions on bacterial fitness29,30. However, studying bacterial growth kinetics in vivo remains intrinsically challenging due to the complexity of the conditions in the hosts, although innovative approaches are being developed31,32. On the other hand, in vitro methods provide a valuable alternative for assessing the impact of specific bacterial and environmental factors on bacterial fitness. Although these in vitro methods have revealed remarkable insights on bacterial fitness, these experiments have mostly yielded strain-specific rather than population-level insights due to the analysis of a small number of isolates from few genetic backgrounds.

Here, we studied the impact of pneumococcal genetics on in vitro growth kinetics to understand the influence of bacterial genetic variation on pneumococcal pathogenesis. Specifically, we performed high-throughput quantification of the in vitro growth kinetics and whole genome sequencing of 348 invasive pneumococcal isolates from the Netherlands to determine the impact of genomic variation on growth kinetics. We conducted growth curve experiments on 348 isolates in blood-like media conditions, and absorbance was monitored over 15 hours. Growth features were determined from these curves using statistical models, and we assessed the impact of capsular serotype and genetic background on the estimated growth features and tested their association with the phylogeny of the isolates. In addition, we used a genome-wide association study (GWAS) approach to identify specific genomic loci associated with the growth features independent of the genetic background. Finally, we calculated the variability in the growth features explained by pneumococcal genetics. Our study suggests that the capsular serotype, genetic background or lineage, and a combination of genetic loci may influence pneumococcal growth kinetics, a fundamental trait potentially influencing nasopharyngeal colonisation, invasive disease pathogenesis, and clinical manifestations.

Results

Population structure of a diverse collection of Dutch invasive pneumococcal isolates and derivation of their growth kinetics

We aimed to understand the impact of the capsular serotype, genetic background, and specific genomic loci, independent of the clonal population structure, on the pneumococcal growth features (Fig. 1 and Supplementary Data 1). Therefore, we investigated the in vitro growth kinetics of 348 pneumococcal isolates collected from patients with invasive pneumococcal diseases via the Pneumococcal Bacteraemia Collection Nijmegen (PBCN) cohort study at two hospitals in Nijmegen, the Netherlands before (2000-2006) and after (2007-2011) the introduction of the seven-valent pneumococcal conjugate vaccine (PCV) [PCV7] into the Dutch paediatric immunisation programme33. Our isolates represent 34 capsular serotypes, 40 pneumococcal lineages, defined based on the global pneumococcal sequence cluster (GPSC) nomenclature34, and 99 sequence types (STs) defined by the pneumococcal multilocus sequence typing (MLST) scheme35. An interactive and annotated maximum likelihood phylogeny of the isolates used in this study is available online at the Microreact platform for open data visualisation and sharing for genomic epidemiology (https://microreact.org/project/wxmdblfgprepwugvbbtaq7). We estimated a Simpson diversity index of 0.932 (95% confidence interval [CI]: 0.924 to 0.940), 0.937 (95% CI: 0.928 to 0.945), and 0.958 (95% CI: 0.949 to 0.966) for the serotypes, lineages, and STs, respectively. The high richness and diversity of pneumococcal serotypes and lineages highlighted a substantial population-level diversity of the study isolates in the Netherlands.

Schematic overview of high-throughput quantification of in vitro pneumococcal growth kinetics and the genome-wide discovery of individual genomic variation influencing in vitro pneumococcal growth.

Population genomic approaches using genome-wide association study (GWAS) offer a robust way to identify genomic variation influencing the growth dynamics of the pneumococcus. First, we collected 348 pneumococcal blood culture isolates that were obtained from 348 patients admitted to the hospital with invasive pneumococcal disease. The isolates were inoculated in a liquid medium in a highly standardised manner followed by measurement of the growth kinetics after every ten minutes for slightly over 15 hours using a microplate reader13. We then fitted a logistic function (see methods) to measure growth features, namely, the average growth rate (r), maximum growth density (Hmax) defined as Lmax, lag phase duration, and maximum growth change (ΔH), defined as the difference between the maximum and minimum density (Lmax-Lmin). Additionally, genomic DNA was extracted from the cultured isolates for whole genome sequencing using the Illumina HiSeq 4000 sequencing platform. The generated sequencing data was used to reconstruct a maximum-likelihood phylogenetic tree of the isolates for downstream analyses, for example, the assessment of phylogenetic signals for the estimated growth features. We also performed de novo assembly of the sequencing reads to generate contigs, which were used to map against the ATCC 700669 pneumococcal reference genome (GenBank accession: NC_011900) to identify consensus single nucleotide polymorphisms (SNPs). We also generated presence, and absence patterns for unitigs or variable length k-mer sequences based on the draft assembled contigs. The SNPs were used to generate the phylogeny of the isolates while the unitigs were used in a GWAS to identify variants associated with the estimated pneumococcal growth features. The images were created with BioRender (https://www.biorender.com/).

Capsular serotype and genetic background significantly impact pneumococcal growth kinetics

To explore the impact of capsular serotype and genetic background or lineage on the pneumococcal growth features estimated from a fitted growth curve, namely lag-phase duration, maximum growth density (Hmax), maximum growth change (ΔH), and average growth rate (r) (Fig. 1, 2, and Supplementary Data 1), we assessed the relationship between these growth features and serotypes and lineages associated with the study isolates (Fig. 3a). In general, we found significant variability of the growth features across serotypes: average growth rate, maximum growth density, maximum growth change, and lag phase duration (P<0.0001; Kruskal-Wallis test) consistent with previous studies14,15 (Fig. 3b-e and Supplementary Data 2). We observed similar patterns across different pneumococcal lineages (Fig. 3f-i and Supplementary Data 2).

Growth curves for selected serotypes show the in vitro pneumococcal growth dynamics and estimated growth features inferred by mathematical modelling.

The growth curves above show the optical density at 620 nm monitored over 15 hours for an example dataset of 20 pneumococcal isolates representing different serotypes and sequence types (STs) based on the pneumococcal MLST scheme. We used the nonlinear optimisation algorithm to identify a combination of parameters with the best fit to a logistic curve based on the lowest residual sum of squares for the predicted value based on the model relative to the observed growth data (see methods). We fitted separate curves based on a subset of points for each replicate up to the end of the stationary phase or end of the experiment, whichever came first. We did this to get a better fit for the logistic curve during the growth phase, as it does not allow for negative growth, which happened after the stationary phase. We fitted the curves to infer the growth features for each of the six replicates. The averaged estimates for each replicate were then used for the downstream analyses. Additional plots showing fitted curves for all the isolates are shown in Supplementary Fig. 1.

The in vitro pneumococcal growth features show variable patterns dependent on the capsular serotype and genetic background or lineage.

(a) Chord diagram showing the association between pneumococcal capsular serotypes and lineages for the Dutch isolates included in this study. The serotypes were determined using in silico approaches based on the sequencing data using SeroBA85, while the lineages were defined using the PopPUNK framework86 based on the global pneumococcal sequencing cluster (GPSC) lineage nomenclature34. The thickness of the connecting lines between serotypes and lineages in the chord diagram represents the percentage of isolates belonging to each serotype and lineage. Due to capsule switching, some serotypes were associated with multiple lineages, and similarly, some lineages were associated with multiple serotypes. The violin plot shows the distribution of the estimated average growth rate across different serotypes. The distribution of the estimated growth features varied across serotypes with at least three isolates, as shown by the violin plots for the (b) average growth rate, (c) maximum growth change, (d) maximum growth change across serotypes, (e) lag phase duration growth features. Similarly, the distribution of the estimated growth features varied across lineages with at least three isolates, as shown by the violin plots for the (f) average growth rate, (g) maximum growth change, (h) maximum growth density, and (i) lag phase duration growth features. A comparison of the growth features between two serotypes belonging to the same lineages with at least three isolates is shown for the (j) growth rate, (k) maximum growth, (l) maximum growth change, and (m) the lag phase duration features. A comparison of the growth features between two lineages expressing the same serotype containing at least three isolates are shown for the (n) growth rates, (o) maximum growth, (p) maximum growth change, and (q) the lag phase features. Additional information regarding the isolates and their estimated growth features or parameters are available in Supplementary Data 2.

Recombination and horizontal gene transfer may lead to the replacement of part or the entire capsule biosynthesis locus resulting in the production of a polysaccharide capsule distinct from the original one, a phenomenon called capsule-switching36. Consequently, some lineages are associated with multiple serotypes, and similarly, a single serotype can be expressed in distinct genetic backgrounds37. Examples of lineages containing multiple serotypes included GPSC3, GPSC4, and GPSC7, and likewise, serotypes, including 3, 19A, and 19F, were expressed in multiple genetic backgrounds, which also express other serotypes (Fig. 3a). Overall, the average growth rate for serotypes expressed in different genetic backgrounds broadly showed similar patterns. For example, lineages expressing serotype 14 had consistently higher average growth rates. In contrast, lineages expressing serotype 3 showed lower rates, consistent with the notion that capsule production interferes with pneumococcal growth14 (Fig. 3b). Therefore, we investigated the impact of the capsular serotype and genetic background on the estimated pneumococcal growth features.

First, we identified lineages that expressed multiple serotypes and serotypes found across multiple lineages. To ensure robust statistical comparison, we only focused on serotype-lineage combinations with at least three sequenced isolates in our study. We found two lineages, with enough sequenced isolates for comparison, GPSC3 and GPSC7, which expressed multiple serotypes (Fig. 3j and Supplementary Data 2). We noted higher growth rates for serotype 33F than serotype 8 (P=0.0007; Kruskal-Wallis test).

In contrast, we found no statistically significant differences in the growth rates between serotypes 23A and 23F in lineage GPSC7 (P=0.2568; Kruskal-Wallis test). We speculated that the absence of statistically significant differences between serotypes 23A and 23F reflected their presence in the same serogroup, therefore, had similar capsule properties. Interestingly, we observed similar patterns for the maximum growth density, maximum growth change, and lag phase duration features between serotypes on the GPSC3 and GPSC7 lineages (Fig. 3k-m and Supplementary Data 2). Next, we investigated the growth rates of the same serotypes expressed in distinct genetic backgrounds. Although some serotypes showed similar growth rates across distinct genetic backgrounds, such as serotype 3 (P=0.4599; Kruskal-Wallis test), others showed different growth rates across different lineages, as seen with serotypes 14 (P=0.0006; Kruskal-Wallis test) and 19F (P=0.0105; Kruskal-Wallis test). We noted similar patterns for the other growth features: maximum growth density, maximum growth change, and lag phase duration (Fig. 3n-q, Supplementary Data 2).

These findings demonstrate that the capsular serotype and genetic background significantly impact pneumococcal growth kinetics. The resultant growth features likely depend on the interaction between these factors, whereby some combinations may result in more pronounced growth effects than others.

Pneumococcal growth kinetics show heritability and correlation with the phylogeny of the isolates

Based on the findings that both capsular serotype and genetic background are associated with growth kinetics, we hypothesised that the estimated pneumococcal growth features exhibit a high phylogenetic signal. We assessed the phylogenetic signal for each growth feature using Pagel’s λ statistic38, which evaluates the correlation between given traits and the phylogenetic tree of the isolates, with values close to zero and one indicating weaker and stronger phylogenetic signals, respectively. By reconstructing the estimated growth features on the internal nodes of the phylogeny, we found high but variable associations between the mapped features and different genetic backgrounds, which suggested potentially strong phylogenetic signals (Fig. 4, Supplementary Fig. 2 and 3, and Supplementary Data 3).

Ancestral reconstruction reveals a strong correlation between the estimated in vitro pneumococcal growth features and the whole-genome phylogeny of the isolates.

The evolution of the normalised estimated continuous growth features, namely average growth rate, maximum growth change, maximum growth density, and lag phase duration, were assessed by estimating the states of the internal nodes of the maximum likelihood phylogenetic tree of the isolates using a maximum likelihood ancestral reconstruction approach92. The length of the scale bar shows the number of nucleotide substitutions per site, while the colours represent the lowest and highest unnormalised growth features. The length of the scale bar shows the number of nucleotide substitutions per site, while the colours represent the lowest and highest unnormalised growth features. A phylogenetic tree showing bootstrap support values on the internal nodes of the tree is provided in Supplementary Data 3.

The amount of the phylogenetic signals based on the normal transformed growth features was highest for the average growth rate (Pagel’s λ=0.72, 95% CI: 0.61 to 0.81; P<2.08×10-64), maximum growth change (Pagel’s λ=0.60, 95% CI: 0.40 to 0.76; P<1.53×10-16), and maximum growth density (Pagel’s λ=0.73, 95% CI: 0.57 to 0.84; P<4.18×10-26), while the lag phase duration showed a lower phylogenetic signal (Pagel’s λ=0.30, 95% CI: 0.17 to 0.47; P<5.42×10-16) (Fig. 5a). These findings suggest that there is a heterogeneous correlation of the growth features with the phylogeny.

To quantify the variability in the pneumococcal growth kinetics explained by the genotype, we calculated the narrow-sense heritability (h2) for the four normal-transformed estimated growth features and the maximum growth rate from Arends et al13. We found strong, though not equal, evidence for a genetic basis for all the estimated growth features: average growth rate (h2=0.76, 95% CI: 0.68 to 0.85), maximum growth change (h2=0.80, 95% CI: 0.71 to 0.89), maximum growth density (h2=0.88, 95% CI: 0.83 to 0.94), and lag phase duration (h2=0.29, 95% CI: 0.14 to 0.43) (Fig. 5b, Supplementary Data 3). To further understand the impact of the capsular serotype and genetic background, we fitted a linear mixed effects model for each growth parameter with serotype and lineage (defined by GPSCs34) as random effects to estimate the amount of variance of the growth features explained by the serotype (h2serotype) and lineage or GPSC (h2GPSC) calculated as the ratio of the variance of the serotype and lineage components relative to the total variance, including the residual or environmental effects. We calculated h2serotype=0.47 and h2GPSC=0.11 for the average growth rate, h2serotype=0.13 and h2GPSC=0.28 for the maximum growth change, h2serotype=0.29 and h2GPSC=0.46 for the maximum growth, h2serotype=0.08 and h2GPSC=0.39 for the lag phase duration feature.

The estimated in vitro pneumococcal growth features show high phylogenetic signals and narrow-sense heritability.

(a) Bar plot showing the phylogenetic signal of the growth features using Pagel’s λ statistic38. Pagel’s λ measures the correlated evolution of the traits with the phylogenetic tree of the isolates. Pagel’s λ estimates close to zero indicate no phylogenetic signal, i.e., the growth features were independent of the phylogeny, while values close to one show a strong phylogenetic signal. The average growth rate, maximum growth change, and maximum growth density showed the highest phylogenetic signal, while the lag phase duration feature showed an intermediate signal. In contrast, the maximum growth rate defined by Arends et al.13 had the lowest phylogenetic signal, i.e., evolved more independently of the phylogeny. (b) Bar plot showing the narrow-sense heritability (h2) of the pneumococcal growth features based on unitig sequences. The narrow-sense heritability corresponds to the amount of variability in the traits, i.e., growth features, explained by the pneumococcal genetic variation. The heritability is the proportion of phenotypic variation explained by the bacterial genetics (PVE) parameter using the linear mixed effects model implemented in GEMMA96. A heritability value close to zero implies low heritability or minimal contribution of genetics to the variability of the traits. In contrast, values close to one suggest a strong influence of genetics on growth features. There was a substantial impact of pneumococcal genetics on all the growth features, with the highest influence seen for the average growth rate, maximum growth change, and maximum growth density, with a slightly lower contribution for the lag phase duration and the maximum growth rate measured by Arends et al.13. All the error bars represent the 95% confidence intervals. Additional information is provided in Supplementary Data 3.

Altogether, these findings demonstrate a substantial impact of the serotype and lineage on the variation of the pneumococcal growth features, whereby most serotypes tend to have similar characteristics regardless of the genetic background. Similarly, some genetic backgrounds tend to show similar phenotypic patterns irrespective of the serotypes, but the serotype appears to have a more significant influence on the growth characteristics.

No individual genomic loci are associated with pneumococcal growth features

Having found moderate to high narrow-sense heritability of the growth features, we then undertook a GWAS analysis to identify specific genomic loci associated with these features independent of the strains’ genetic background using a linear mixed-effects model. We performed separate GWAS analyses of the maximum growth change, maximum growth density, maximum growth change, and lag phase duration phenotypes derived from the growth curve. We conducted the GWAS using variable-length k-mer sequences or unitigs to capture genetic variation in the pneumococcal genomes at different levels, including SNPs, multi-allelic sites, and insertions and deletions39,40. We found no individual unitigs associated with the derived growth feature phenotypes (Fig. 6a, Supplementary Data 4). Furthermore, repeating the GWAS analyses adjusting for the serotype revealed no statistically significant hits associated with the derived growth phenotypes, suggesting that the variants in the pneumococcal strains are correlated with the growth kinetics independent of the serotype (Fig. 6b, Supplementary Data 4). Overall, the observed and expected statistical significance values in the QQ-plots for our study revealed no apparent issues with the population structure of the isolates, which is a significant confounder in similar studies41 (Fig. 6c, d). Overall, these findings suggest that no individual genomic variation influences the pneumococcal growth rates, suggesting that multiple genetic loci and lineage effects modulate pneumococcal growth kinetics.

Pneumococcal genetic variation associated with the in vitro growth features or parameters of the isolates.

Manhattan plot showing the statistical significance of the hits from the GWAS analysis based on unitigs (a) without adjusting for the serotype and (b) after adjusting for the serotype. The statistical significance (P-value) were inferred based on the likelihood ratio test using Pyseer94 mapped against the ATCC 700669 pneumococcal reference genome (GenBank accession: NC_011900101) to identify their location in the pneumococcal genome. The dots in the Manhattan plots with different colours represent GWAS for different growth features, as shown in the key. The red dotted line designates the genome-wide threshold for considering a variant as significantly associated, statistically, with the growth features. The blue dotted line represents the threshold for classifying variants as suggestively associated with the growth features. (c) Quantile–quantile plots showing the relationship between the observed statistical significance and the expected statistical significance for the GWAS of growth features not adjusting for the serotype. (d) Quantile–quantile (QQ) plots showing the relationship between the observed statistical significance and the expected statistical significance for the GWAS of growth features after adjusting for the capsular serotype.

Average in vitro growth rates are not strongly correlated with a serotype’s invasiveness or its prevalence in carriage or disease

Previous studies have elucidated that bacterial growth rate may influence colonisation and disease16, including systemic dissemination during infection26 and disease manifestations13,27. Pneumococcal strains that are prolific growers may display better survival and replication in systemic tissues than slow growers without vaccine-induced non-serotype-specific immunity. We used the isolates collected before the introduction of the PCV7 vaccine in the Netherlands42 to assess the association between in vitro growth rate of the isolates and their population-level characteristics: prevalence in invasive disease and nasopharyngeal carriage and invasiveness, i.e., odds of being isolated in invasive disease compared to the carriage state. Our first hypothesis was that the average in vitro growth rate for each of the pneumococcal serotypes is positively associated with the prevalence of serotypes in invasive diseases. Our analysis revealed no evidence of an association between the prevalence of the serotypes in invasive disease and the in vitro growth rate (Spearman correlation: -0.30, P=0.0856) (Fig. 7a, Supplementary Data 4).

Association between average in vitro pneumococcal growth rate with the prevalence of serotypes in invasive disease, nasopharyngeal carriage, and invasiveness.

(a) Scatter plot showing the relationship between the prevalence of serotypes among invasive disease isolates collected before the introduction of PCV7 in the Netherlands and the average in vitro growth rate (r) of the isolates inferred as shown in Fig. 1. A Spearman correlation test showed no evidence of an association between the growth rate and prevalence of serotypes among the invasive disease isolates. (b) Scatter plot showing the relationship between the inferred prevalence of serotypes in the nasopharyngeal carriage before the introduction of PCV7 in the Netherlands and the inferred average in vitro growth rate. Due to the unavailability of the nasopharyngeal carriage data, we imputed the prevalence of the serotypes in carriage by dividing the prevalence in invasive disease by the average invasiveness of each serotype. The invasiveness index was calculated based on the meta-analysis of data collected before PCV7 introduction across multiple countries (see methods). A Spearman correlation test of the growth rate and prevalence of serotypes in carriage suggested a negative association between the variables. The association remained even after excluding serotype 3, which seemed like an outlier. (c) The relationship between the invasiveness and the average growth rates of each serotype from the pre-vaccination data. A Spearman correlation test of the growth rates and invasiveness suggested no association between the variables. All the error bars represent the 95% confidence intervals. Additional information is provided in Supplementary Data 4.

Second, we hypothesized that the average in vitro growth rate of the serotypes is positively associated with the prevalence of the serotypes in the nasopharyngeal carriage as fitter strains would grow faster than less fit strains. However, contrary to our hypothesis, we found a negative association between the prevalence of the serotypes in carriage and the in vitro growth rate (Spearman correlation=-0.53, P=0.0098) (Fig. 7b, Supplementary Data 4). We repeated the analysis after excluding serotype 3 and found no evidence of an association between growth rate and carriage prevalence (Spearman correlation=0.38, P=0.0925) which suggests that serotype 3 alone was driving the previously observed association. Finally, we tested the hypothesis that the growth rate is positively associated with the invasiveness of the serotypes. We found no association between the invasiveness and the in vitro growth rates of the serotypes (Spearman correlation: 0.35, P=0.1024) (Fig. 7c and Supplementary Data 4). These findings suggest that growth rates alone may not be sufficient to explain the prevalence and invasiveness of pneumococcal serotypes, which suggests the influence of other pneumococcal, microbiome (e.g., viral infection), and host factors in determining pneumococcal disease susceptibility.

Discussion

Bacterial growth kinetics impact several traits, including gene and protein expression17,18, antibiotic accumulation and efficacy23–25, cell surface adhesion19, intra- and inter-specific competition20, response to environmental conditions14–16, plasmid replication21, bacteriophage dynamics22, colonisation and disease pathogenesis12, including clinical manifestation12–14. Whilst previous studies have attempted to understand bacterial, host, and environmental factors influencing bacterial growth, a major drawback has been a narrow focus on a small subset of strains43. Therefore, the insights gained from these studies have mostly been for specific strains, as they ignored potential variability across different genetic backgrounds. These population-level studies are critical for genetically diverse bacterial pathogens with highly structured populations34, substantial antigenic diversity44, and recombination rates, such as the pneumococcus8,36,45. Here, we examine the association between pneumococcal genetics and growth kinetics using a combination of high-throughput quantification of in vitro pneumococcal growth, whole-genome sequencing, and population genomic analyses employing a bacterial GWAS approach. Notably, our findings reveal a strong influence of the capsular serotype compared to the genetic background, and the absence of specific individual genomic loci modulating pneumococcal growth kinetics, particularly the in vitro growth rate. These findings may partially explain why some serotypes and lineages colonise and cause invasive diseases better than others and may have implications for developing measures to prevent and control pneumococcal infections.

Our findings suggest that the capsular serotype and genetic background influence pneumococcal growth kinetics. The findings are generally consistent with results from experimental studies by others13–15, which used capsule-switched strains to exclude the impact of the genetic background and showed an alternation of growth dynamics due to the capsule, although there may be potential differences in the growth media and conditions. Specifically, encapsulated strains showed higher growth rates than unencapsulated strains in nutrient-rich conditions14. These findings suggest that the capsule is crucial for pneumococcal growth, but its effect depends on nutrient availability – promoting and interfering with growth when nutrients are abundant and limited, respectively14,27. A recent study by Tóthpál et al.15 suggested that the serotype partially explains pneumococcal growth and suggested the genetic backgrounds, which vary in their ability to colonise the nasopharynx and cause invasive diseases4,46–48, may also play a role. However, the authors did not fully quantify the impact of the genetic background since the isolates were not sequenced15. Our study confirms the strong influence of both the pneumococcal capsular serotype and genetic background or lineage on the growth kinetics of the isolates. The influence of the serotype appeared to be higher than the genetic background for the average growth rate, while the other growth features appeared to show an opposite trend. These findings suggest that the respective influence of the capsular serotype and lineage varies by specific derived pneumococcal growth features.

Interestingly, although our findings revealed that the capsular serotype and genetic background influence pneumococcal growth kinetics, statistical analysis showed that the variance attributed to the capsular serotype was higher than the genetic background. Specifically, the capsular serotype exhibits a more pronounced growth rate, maximum growth change, and maximum growth density than the genetic background. In contrast, the genetic background had a higher effect on the lag phase duration than the serotype. We acknowledge that our comparison of growth features in the same genetic background could also mean a large amount of genetic variation for some of the lineages, some of which could be responsible for changes in growth. These findings illustrate the importance of the capsular serotype and genetic background on pneumococcal growth kinetics, which may partially explain their variable prevalence in carriage and disease.

Previous studies have revealed individual bacterial genomic loci that impact other phenotypes, including pathogenicity49–57, virulence58–61, host, niche, and environmental adaptation52,54,62, conjugation63, and antimicrobial resistance64–71, independent of the genetic background. Our GWAS findings revealed no elevated signals in specific genomic loci, suggesting that individual variants may not significantly influence pneumococcal growth kinetics independent of the genetic background and serotype. Previous studies have shown that mutations in other capsule biosynthesis genes (wchA or cpsE) encoding a protein that links activated glucose phosphates to lipid carriers44 also influence the growth dynamics of the pneumococcus43. However, our results are contrary to our speculation that genetic variation in the capsule polysaccharide synthesis genes would be strongly associated with pneumococcal growth features. These findings suggest that specific genomic loci in the pneumococcal genome do not strongly influence pneumococcal growth kinetics, hinting at the potential impact of the overall genetic background and several loci possibly with small effects individually. While it’s possible that such follow-up studies are needed to provide additional support for the importance of these variants on pneumococcal growth kinetics.

In contrast to findings in a previous study14, we found no robust evidence for the association between the in vitro growth rates and the population-level prevalence of the serotypes in nasopharyngeal carriage. However, this may have occurred as a result of a limitation of our study, since we studied a collection of randomly sampled invasive isolates, and the growth experiments were conducted in conditions similar to that of blood, while the Hathaway et al. study14 used a non-random selection of both carriage and disease isolates. It is likely that isolates have adaptations to the environmental conditions in which they were isolated, and so in vitro growth characteristics for any given isolate may be different under experimental conditions simulating different niches. In addition, the in vitro growth environment cannot fully represent in vivo conditions and fails to consider the impact of selective pressures coming from the serotype-specific responses (antibody) or non-specific responses (phagocytic) and potential interactions with other pathogens. Presumably, the lack of association between the invasiveness and prevalence of serotypes in invasive disease and carriage, and the growth rates could reflect that the conditions in the in vitro growth environment do not fully represent in vivo conditions. Collectively, these findings may suggest that these and other factors may substantially influence the complex pneumococcal disease pathogenesis in vivo.

The strength of this study is the analysis of an extensive collection of well-phenotyped natural isolates and the combination of phenotyping, and estimation of growth features using statistical models and statistical genetics to determine the association between the growth features and genetic variation across the entire genome. As previously noted by Tothpal et al.15, the culturing history of the isolates may impact the growth kinetics. In our study, we consistently cultured all the isolates after isolation from patients using standardised culturing procedures (single assay, identical initial concentration, environmental conditions or temperature, and equipment to measure the optical density and repeat measurements to spot technical variability13). Compared to the present study, future studies would benefit from larger sample sizes and may find more results as we only explained some of the heritability of the growth features. We also acknowledge that the methods used to estimate heritability may also sometimes perform poorlyon some bacteria species, therefore, more robust methods are needed to further validate our findings72. Because we conducted our experiments for a relatively short duration (∼15 hours), we could not robustly characterise the kinetics at the stationary phase due to incomplete data or the observed high variability of its kinetics. Therefore, we recommend increasing the culturing time from 15 to >24 hours in follow-up studies, as done elsewhere14,15, to characterise the stationary phase kinetics of the cultured isolates robustly. Furthermore, since the pneumococcus invades different niches and systemic tissues with varying conditions, including oxygen concentration and temperature12,14–16, future studies should investigate the effect of these environmental conditions on pneumococcal growth and identify pneumococcal genomic variants associated with growth kinetics for each condition. Considering that our dataset contained only a third of the >100 known pneumococcal serotypes globally3; due to the geographical variation in the distribution of pneumococcal serotypes73, future studies should investigate all serotypes to provide further insights on the impact of different serotypes in pneumococcal growth kinetics. Additionally, considering that we generated the phylogeny of the entire pneumococcal species rather than isolates belonging to a single strain or lineage, as defined by several approaches34,74,75, we did not remove recombinations using widely used methods designed for clonal bacterial populations, such as Gubbins76. Therefore, there is a possibility that some branch lengths in the phylogeny used for the ancestral trait reconstruction analysis may be less accurate although the topology is usually robust to recombination77. Furthermore, there may be other variants that potentially influence the pneumococcal growth kinetics which are regarded as not statistically significant in this study based on our conservative genome-wide significance threshold. As more and more similar studies are conducted, particularly those utilising much larger datasets, potential variants associated with pneumococcal growth kinetics, may be discovered.

Our analysis of population-level genomic and high-throughput in vitro pneumococcal growth data suggests that the capsular serotype and lineage influence pneumococcal growth. However, although some growth features are highly heritable, there appears to be no specific genetic variants which strongly influence these phenotypes suggesting that a combination of genomic loci with small effects individually or the entire genetic background may play a greater role in pneumococcal growth kinetics. Overall, our study provides a proof-of-concept for combining phenotyping and population genomics to understand the genetic basis for pneumococcal growth, an approach which can be applied to study other bacteria of clinical significance and those used in bioengineering, genetics, and the food industry78–80. Ultimately, we anticipate that applying our approach will have implications for devising strategies to reduce life-threatening bacterial infections and improve food production and safety.

Methods

Ethics statement

The isolates used in this work were collected through an observational cohort study Pneumococcal Bacteraemia Collection Nijmegen (PBCN)33. The non-applicability of the Medical Research Involving Human Subjects Act (WMO) on the study design was confirmed by the Regional Ethics Committee, and the local medical ethics committees of the participating hospitals in the Netherlands approved the study procedures.

Sample characteristics and whole-genome sequencing

Three hundred and forty-eight (n=348) pneumococcal isolates collected from patients with invasive pneumococcal diseases through the PBCN observational cohort study33 were available for the study. We included a single isolate per patient as invasive pneumococcal infections are typically clonal, so a single isolate is sufficient. The consecutive blood isolates were collected at two hospitals in Nijmegen, the Netherlands, before (2000-2006) and after (2007-2011) the introduction of the PCV7 vaccine into the Dutch paediatric immunisation programme33. Most of the bacteremia cases (97%) concerned adults. Additional information on the PBCN study and collected clinical variables have been previously described33,81. To generate whole genome sequences, we sequenced a single isolate per patient using an Illumina HiSeq sequencer (Illumina, San Diego, CA, USA). We then assembled the generated reads using SPAdes genome assembler (version 3.14.0)82, and evaluated the draft assemblies using assembly-stats (version: 1.0.1) (https://github.com/sanger-pathogens/assembly-stats). In addition, we used a combination of metrics, including genome size and the number of contigs, to assess the quality of the draft assemblies for inclusion in the analysis.

Measurement of in vitro pneumococcal growth kinetics and derivation of growth features

We measured the pneumococcal growth kinetics for slightly over 15 hours as described by Arends et al.13. In short, we incubated a frozen aliquot of standardised pneumococcal inoculum in 1.5 ml of blood-like medium (50% M17, 50% casamino acids tryptone [CAT] medium, 0.25% glucose). We measured growth kinetics at 37°C and 5% CO2 every 10 min for over 15 hours at OD620 with use of a humidity cassette by microplate reader (Spark 10M, Tecan, Switzerland). We added 15 μL of inoculum was added to 1.5 mL of prewarmed rich growth medium (45% M17, 45% CAT, 0.225% glucose, 10% fetal calf serum [Greiner Bio-one], 26 U/mL catalase [Sigma-Aldrich C1345]) in each well of a sterile flat-bottomed 48-well plate (Nunclon Surface, Nunc, Denmark). The growth medium was supplemented with catalase. We chose these nutrient-rich conditions to bring forward differences in the growth features of the pneumococcus. We performed six repeat measurements on three separate days for each isolate. To understand the growth kinetics of the invasive Dutch pneumococcal isolates, we fitted a modified logistic growth curve function, i.e., L0 + Lmax/(1 + e−r(t−t0)), separately to the growth data for each isolate for each replicate, using the “optim” function in stats (version 4.3.2) R package (https://www.R-project.org/). We then averaged the inferred values for each replicate to obtain average estimates for each parameter. The parameter L0 captures the initial cell density measured as absorbance or optical density at a wavelength of 620 nm (OD620); Lmax is the maximum cell density; t0 is the initial time, and r is the average growth rate. We fitted two curves in total for each isolate based on the averaged data from the six replicates using the limited-memory Broyden-Fletcher-Goldfarb-Shanno (L-BFGS-B) optimisation algorithm by minimising the residual sum of squares83. We fitted the first curve on all the data points and the second curve on the data measured until the start of the end of the stationary phase or the beginning of the death phase, whichever came earlier. We defined the end of the stationary phase or the start of the death phase as the time when the optical cell density decreased by >5% from the maximum cell density. The curve fitted based on the data measurements collected up to the end of the stationary phase or death phase, or the end of the experiment, if none of these phases had been reached, captured the growth dynamics better than the curve fitted using all the data points to accurately estimate the growth rate and the maximum cell density corresponding to the peak of the pneumococcal growth curve, especially for growth curves with observably short stationary phases and rapid autolysis. Therefore, parameters for the downstream analyses were inferred from the curves fitted to the reduced dataset.

We determined four growth phenotypes for each pneumococcal isolate estimated from the fitted growth curve, namely the lag-phase duration, maximum growth density (Hmax), i.e., captured by parameter Lmax, maximum growth change (ΔH) corresponding to the difference of the maximum and minimum density (Lmax-L0), and average growth rate (r). We defined the lag-phase duration as the time when the optical cell density was at least 1.5-fold higher than the starting initial cell density, as described by Arends and colleagues13. Due to the challenges of characterising the kinetics at the stationary or death phase, partly due to the high strain-to-strain variability or short study period, we did not derive the growth features capturing these phases, including the duration of the stationary and death phases. The logistic growth curves showed an excellent fit for the data, which provided an opportunity to investigate the impact of capsular serotype and genetic background or lineages and specific genomic loci, independent of the genetic background, on the estimated growth parameters or features.

Molecular typing of sequenced isolates

Molecular typing of the isolates to determine capsular serotypes and sequence types (STs) was done in silico using Pathogenwatch (https://pathogen.watch/), a global platform for genomic surveillance84. We determined the capsular serotypes and STs using SeroBA (version v1.0.1)85 and the multilocus sequence typing (MLST) scheme for the pneumococcus35, respectively. Pneumococcal lineages were defined based on the global pneumococcal sequence cluster (GPSC) nomenclature34 using PopPUNK (version 1.1.0)86. In addition, we checked the species of each sequenced isolate using Speciator (version 3.0.1) in Pathogenwatch.

Construction of multiple sequence alignments and phylogenetic tree

The generated draft assemblies for each isolate were mapped to the pneumococcal ATCC 700669 reference genome (GenBank accession: NC_011900) using Snippy (version 4.6.0) (https://github.com/tseemann/snippy). We identified 155,305 SNPs in the generated alignment using SNP-sites (version 2.3.2)87 and then used them to construct a maximum-likelihood phylogeny using IQ-TREE (version 2.0.3)88. We used the alignment of variable sites by mapping to a reference genome to generate a maximum-likelihood phylogeny using IQ-TREE as it has been to produce more accurate phylogenies89. The general time reversible (GTR) nucleotide substitution model and Gamma heterogeneity of the substitution rates among sites. We assessed branch support with ultrafast bootstrap approximation using “-B 1000” option90. We optimised bootstrapped trees by nearest neighbour interchange on bootstrap alignment by specifying the “--bnni” option. We rooted the phylogenetic tree of the pneumococcal isolates using Streptococcus oralis as an outgroup (not shown in the phylogeny). An annotated and interactive version of the phylogeny is available on the Microreact web tool91 (https://microreact.org/project/wxmdblfgprepwugvbbtaq7).

Quantifying the phylogenetic signal

To assess the correlation between the estimated growth features and pneumococcal phylogeny, we used models of continuous character evolution to estimate the phylogenetic signal using Pagel’s λ38 implemented in the “phylosig” function in phytools package (version 2.1.1)92. Pagel’s λ transforms the internal branches relative to the terminal branches and generates values ranging from zero to one, with smaller values indicating no phylogenetic signal or independent evolution of the phenotype from the phylogeny. In comparison, higher values close to one show a high phylogenetic signal or a decreased phenotype variability among genetically similar taxa at the tips of the phylogeny. First, we transformed the estimated growth features to a normal distribution using a rank-based inverse normal transform implemented in the RNOmni package (version 1.0.1.2) (https://cran.r-project.org/web/packages/RNOmni/). Next, we mapped the estimated continuous growth feature values onto the phylogeny using fast estimation of maximum likelihood for ancestral states92.

Bacterial GWAS

We identified the presence and absence patterns of unitig sequences To capture genetic variation in the pneumococcal genomes included in this study for the GWAS analysis. We identified unitig sequences from 31bp k-mers using Bifrost (version 1.0.6)93. First, we inferred unitig sequences from a compacted De Bruijn graph constructed using k-mers of length 31 using draft assemblies of all the pneumococcal isolates included in this study. We then generated De Bruijn graphs for each isolate and queried them with the unitig sequences found in the entire dataset to identify the presence and absence of each unitig found in each sequenced genome. We considered a unitig present when 100% of the k-mer was presented in the De Bruijn graph. We combined the presence and absence patterns of the unitig sequences from all the isolates to construct a single matrix.

Since the linear mixed effects models require the tested continuous phenotype to be normally distributed, we used the normalised growth features after applying the rank-based inverse normal transformation using the RNOmni package (version 1.0.1.2) (https://cran.r-project.org/web/packages/RNOmni/). We performed GWAS analysis of the pneumococcal growth features using robust linear mixed effects models implemented in Pyseer (version 1.3.7-dev)94. A kinship matrix specifying the variances and covariances of the random effects for each isolate was generated using the “similarity_pyseer” script in Pyseer. Using this kinship matrix, Pyseer fits an unobserved random effect for each isolate to account for the clonal population structure in the GWAS. We excluded unitigs with <5% minimum allele frequency and >95% maximum allele frequency. The inferred statistical significance for each variant was adjusted to correct for multiple testing using the Bonferroni correction. We estimated the total number of unique unitig presence and absence patterns in the dataset using a script included in Pyseer (https://github.com/mgalardini/pyseer/blob/master/scripts/count_patterns.py), and used the obtained value when adjusting for multiple testing using the Bonferroni correction approach. Therefore, we defined the genome-wide statistical significance threshold as P <2.58×10-07, i.e., α/G, where α=0.05 and G=194,053 is the number of unique unitig presence and absence patterns while the suggestive threshold as P<5.15×10-06, whereby α=1. To check potential issues arising from the population structure, we generated quantile-quantile (QQ) plots from the output of GWAS analysis for each growth feature to compare the expected and observed P-values using qqman (version 0.1.7)95. The overall proportion of phenotypic variability explained by variation in the genome (narrow-sense heritability [h2]) was estimated using GEMMA (version 0.98.1)96 and Pyseer. The input files for GEMMA were prepared based on the unitig presence and absence data using PLINK (version 1.90b4) 97. To identify variants associated with the growth features independently of the serotype, we repeated the GWAS by including the serotype as a covariate.

Annotation of variants identified by the GWAS analysis

To annotate the unitig sequences with statistical significance values below the suggestive and genome-wide significance threshold, we used a custom BioPython 98 wrapper script for BLASTN (version 2.5.0+)99 to identify genomic features in the ATCC 700669 reference genome (GenBank accession: NC_011900) and other complete pneumococcal genomes downloaded from GenBank99. The GWAS results were summarised and visualised using Manhattan plots using R (version 4.0.3) [https://www.R-project.org/].

Other statistical analysis

We compared the growth features between serotypes and lineages using the Kruskal-Wallis non-parametric test. The distribution of the growth features was plotted using violin plots using the vioplot package [version 0.4.0] (https://CRAN.R-project.org/package=vioplot). To assess the contribution of serotype and lineage to the variability in the pneumococcal growth features, we used linear mixed-effects models implemented in the lme4 package (version 1.1.35.1)100 were used, with each growth feature specified as the outcome and the serotype and lineage as random effects. The statistical significance for the linear mixed model was inferred using the lmerTest package [version 3.1.3] (https://CRAN.R-project.org/package=lmerTest). To assess the contribution of the serotype and lineage to the variability of each growth feature, we calculated the proportion of variance for the serotype and lineage components to the total variance. We calculated the average invasiveness of the serotypes based on a meta-analysis of data collected in different countries in infants before the introduction of the PCVs47 using the random-effects model implemented in the metafor package [version 2.4.0] (https://CRAN.R-project.org/package=metafor We defined invasiveness as the odds ratio of detecting a pneumococcal serotype among isolates causing invasive disease compared to nasopharyngeal carriage. We added a value of one to cells with zero counts. Due to the unavailability of the nasopharyngeal carriage data, we imputed the prevalence of the serotypes in the carriage by dividing the prevalence in invasive disease by the average invasiveness of each serotype. Next, we used a Spearman correlation test to assess the association between the mean prevalence of serotypes in the nasopharyngeal carriage and invasive disease and average invasiveness with the average in vitro pneumococcal growth rate.

eLife Assessment

This is an important study that examines the impact of Streptococcus pneumoniae genetics on its in vitro growth kinetics, aiming to identify potential targets for vaccines and therapeutics. The study identified significant variations in growth characteristics among capsular serotypes and lineages, linked to phylogeny and high heritability, but genome-wide association studies did not reveal specific genomic loci associated with growth features independent of the genetic background. The evidence supporting these findings is solid.

Reviewer #1 (Public review):

Summary:

This manuscript uses a diverse isolate collection of Streptococcus pneumoniae from hospital patients in the Netherlands to understand the population-level genetic basis of growth rate variation in this pathogen, which is a key determinant of S. pneumoniae within-host fitness. Previous efforts have studied this phenomenon in strain-specific comparisons, which can lack the statistical power and scope of population-level studies. The authors collected a rigorous set of in vitro growth data for each S. pneumoniae isolate and subsequently paired growth curve analysis with whole-genome analyses to identify how phylogenetics, serotype, and specific genetic loci influence in vitro growth. While there were noticeable correlations between capsular serotype and phylogeny with growth metrics, they did not identify specific loci associated with altered in vitro growth, suggesting that these phenotypes are controlled by the collective effect of the entire genetic background of a strain. This is an important finding that lays the foundation for additional, more highly-powered studies that capture more S. pneumoniae genetic diversity to identify these genetic contributions.

Strengths:

(1) The authors were able to completely control the experimental and genetic analyses to ensure all isolates underwent the same analysis pipeline to enhance the rigor of their findings.

(2) The isolate collection captures an appreciable amount of S. pneumoniae diversity and, importantly, enables disentangling the contributions of the capsule and phylogenetic background to growth rates.

(3) This study provides a population-level, rather than strain-specific, view of how genetic background influences the growth rate in S. pneumoniae. This is an advance over previous studies that have only looked at smaller sets of strains.

(4) The methods used are well-detailed and robust to allow replication and extension of these analyses. Moreover, the manuscript is very well written and includes a thoughtful and thorough discussion of the strengths and limitations of the current study.

Weaknesses:

(1) As acknowledged by the authors, the genetic diversity and sample size of this newly collected isolate set are still limited relative to the known global diversity of S. pneumoniae, which evidently limits the power to detect loci with smaller/combinatorial contributions to growth rate (and ultimately infection).

(2) The in vitro growth data is limited to a single type of rich growth medium, which may not fully reflect the nutritional and/or selective pressures present in the host.

(3) The current study does not use genetic manipulation or in vitro/in vivo infection models to experimentally test whether alteration of growth rates as observed in this study is linked to virulence or successful infection. The availability of a naturally diverse collection with phylogenetic and serotype combinations already identified as interesting by the authors provides a strong rationale for wet-lab studies of these phenotypes.

Reviewer #2 (Public review):

Summary:

The study by Chaguza et al. presents a novel perspective on pneumococcal growth kinetics, suggesting that the overall genetic background of Streptococcus pneumoniae, rather than specific loci, plays a more dominant role in determining growth dynamics. Through a genome-wide association study (GWAS) approach, the authors propose a shift in how we understand growth regulation, differing from earlier findings that pinpointed individual genes, such as wchA or cpsE, as key regulators of growth kinetics. This study highlights the importance of considering the cumulative impact of the entire genetic background rather than focusing solely on individual genetic loci.

The study emphasizes the cumulative effects of genetic variants, each contributing small individual impacts, as the key drivers of pneumococcal growth. This polygenic model moves away from the traditional focus on single-gene influences. Through rigorous statistical analyses, the authors persuasively advocate for a more holistic approach to understanding bacterial growth regulation, highlighting the complex interplay of genetic factors across the entire genome. Their findings open new avenues for investigating the intricate mechanisms underlying bacterial growth and adaptation, providing fresh insights into bacterial pathogenesis.

Strengths:

This study exemplifies a holistic approach to unraveling key factors in bacterial pathogenesis. By analyzing a large dataset of whole-genome sequences and employing robust statistical methodologies, the authors provide strong evidence to support their main findings. Which is a leap forward from previous studies focused on a relatively smaller number of strains. Their integration of genome-wide association studies (GWAS) highlights the cumulative, polygenic influences on pneumococcal growth kinetics, challenging the traditional focus on individual loci. This comprehensive strategy not only advances our understanding of bacterial growth regulation but also establishes a foundation for future research into the genetic underpinnings of bacterial pathogenesis and adaptation. The amount of data generated and corresponding approaches to analyze the data are impressive as well as convincing. The figures are convincing and comprehensible too.

Weaknesses:

Despite the strong outcomes of the GWAS approach, this study leaves room for differing interpretations. A key point of contention lies in the title, which initially gives the impression that the research addresses growth kinetics under both in vitro and in vivo conditions. However, the study is limited to in vitro growth kinetics, with the assumption that these findings are equally applicable to in vivo scenarios-a premise that is not universally valid. To more accurately reflect the study's scope and avoid potential misrepresentation, the title should explicitly specify "in vitro" growth kinetics. This clarification would better align the title with the study's actual focus and findings.

This study suggests that the entire genetic background significantly influences bacterial growth kinetics. However, to transform these predictions into established facts, extensive experimental validation is necessary. This would involve "bench experiments" focusing on generating and studying mutant variants of serotypes or strains with diverse genomic variations, such as targeted deletions. The growth phenotypes of these mutants should be analyzed, complemented by complementation assays to confirm the specific roles of the deleted regions. These efforts would provide critical empirical evidence to support the findings from the GWAS approach and enhance understanding of the genetic basis of bacterial growth kinetics.

In the discussion section, the authors state that "the influence of serotype appeared to be higher than the genetic background for the average growth rate" (lines 296-298). Alongside references 13-15, this emphasizes the important role of capsular variability, which is a key determinant of serotypes, in influencing growth kinetics. However, this raises the question: why isn't a specific locus like cps, which is central to capsule biogenesis, considered a strong influencer of growth kinetics in this study?

One plausible explanation could be the absence of "elevated signals" for cps in the GWAS analysis. GWAS relies on identifying loci with statistically significant associations to phenotypes. The lack of such signals for cps may indicate that its contribution, while biologically important, does not stand out genome-wide. This might be due to the polygenic nature of growth kinetics, where the overall genetic background exerts a cumulative effect, potentially diluting the apparent influence of individual loci like cps in statistical analyses.

Reviewer #3 (Public review):

This study provides insights into the growth kinetics of a diverse collection of Streptococcus pneumoniae, identifying capsule and lineage differences. It was not able to identify any specific loci from the genome-wide association studies (GWAS) that were associated with the growth features. It does provide a useful study linking phenotypic data with large-scale genomic population data. The methods for the large part were appropriately written in sufficient detail, and data analysis was performed with rigour. The interpretation of the results was supported by the data, although some additional explanation of the significance of e.g. ancestral state reconstruction would be useful. Efforts were made to make the underlying data fully accessible to the readers although some of the supplementary material could be formatted and explained a bit better.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

Summary:

This manuscript uses a diverse isolate collection of Streptococcus pneumoniae from hospital patients in the Netherlands to understand the population-level genetic basis of growth rate variation in this pathogen, which is a key determinant of S. pneumoniae within-host fitness. Previous efforts have studied this phenomenon in strain-specific comparisons, which can lack the statistical power and scope of population-level studies. The authors collected a rigorous set of in vitro growth data for each S. pneumoniae isolate and subsequently paired growth curve analysis with whole-genome analyses to identify how phylogenetics, serotype, and specific genetic loci influence in vitro growth. While there were noticeable correlations between capsular serotype and phylogeny with growth metrics, they did not identify specific loci associated with altered in vitro growth, suggesting that these phenotypes are controlled by the collective effect of the entire genetic background of a strain. This is an important finding that lays the foundation for additional, more highly-powered studies that capture more S. pneumoniae genetic diversity to identify these genetic contributions.

Thank you for an excellent summary of our manuscript.

Strengths:

(1) The authors were able to completely control the experimental and genetic analyses to ensure all isolates underwent the same analysis pipeline to enhance the rigor of their findings.

(2) The isolate collection captures an appreciable amount of S. pneumoniae diversity and, importantly, enables disentangling the contributions of the capsule and phylogenetic background to growth rates.

(3) This study provides a population-level, rather than strain-specific, view of how genetic background influences the growth rate in S. pneumoniae. This is an advance over previous studies that have only looked at smaller sets of strains.

(4) The methods used are well-detailed and robust to allow replication and extension of these analyses. Moreover, the manuscript is very well written and includes a thoughtful and thorough discussion of the strengths and limitations of the current study.

Thank you for excellently summarising the strengths of our manuscript.

Weaknesses:

(1) As acknowledged by the authors, the genetic diversity and sample size of this newly collected isolate set are still limited relative to the known global diversity of S. pneumoniae, which evidently limits the power to detect loci with smaller/combinatorial contributions to growth rate (and ultimately infection).

Indeed, while larger pneumococcal datasets exist globally, most of these datasets do not have reliable metadata on in vitro growth rates and other phenotypes, as the intention, for the most part, is to conduct population-level surveillance to track the changes in the serotype distribution to assess the impact of introducing pneumococcal conjugate vaccines. In this study, we adopted a different approach to phenotypically characterising the samples collected from these surveillance studies to understand the genetic features that influence the intrinsic growth characteristics of the isolates. While our dataset size is modest, it exemplifies how we can combine whole-genome sequencing and phenotypic characterisation of bacterial isolates to understand the genetic determinants that may drive intrinsic phenotypic differences between strains.

(2) The in vitro growth data is limited to a single type of rich growth medium, which may not fully reflect the nutritional and/or selective pressures present in the host.

We agree that our study focused on a single type of rich growth medium, which may not fully reflect the nutritional or selective pressures present in the host. The rationale and the representativeness of the selected culture conditions were more extensively discussed in Arends et al. (10.1128/spectrum.00050-22). Considering that this was a proof-of-concept study to assess the feasibility of our approach, future studies by us and others will evaluate the impact of using different media. Besides the media, complementary techniques such as transcriptome sequencing will help uncover additional insights into potential factors that influence differences in pneumococcal growth kinetics.

(3) The current study does not use genetic manipulation or in vitro/in vivo infection models to experimentally test whether alteration of growth rates as observed in this study is linked to virulence or successful infection. The availability of a naturally diverse collection with phylogenetic and serotype combinations already identified as interesting by the authors provides a strong rationale for wet-lab studies of these phenotypes.

We concur that additional genetic manipulation studies to assess the impact of altering growth rates on virulence and infection would have provided further insights. While this was beyond the scope of this study, we plan to conduct follow-up work to assess this using carefully selected strains from our pneumococcal collection. Because our current study demonstrates that genetic determinants of pneumococcal growth features are not simply confined to single loci, such experimental validation would require novel wet-lab approaches that consider epistatic interactions. In addition, in vivo infection models that allow the study of dissemination from the bloodstream are not yet well established.

Reviewer #2 (Public review):

Summary:

The study by Chaguza et al. presents a novel perspective on pneumococcal growth kinetics, suggesting that the overall genetic background of Streptococcus pneumoniae, rather than specific loci, plays a more dominant role in determining growth dynamics. Through a genome-wide association study (GWAS) approach, the authors propose a shift in how we understand growth regulation, differing from earlier findings that pinpointed individual genes, such as wchA or cpsE, as key regulators of growth kinetics. This study highlights the importance of considering the cumulative impact of the entire genetic background rather than focusing solely on individual genetic loci.

The study emphasizes the cumulative effects of genetic variants, each contributing small individual impacts, as the key drivers of pneumococcal growth. This polygenic model moves away from the traditional focus on single-gene influences. Through rigorous statistical analyses, the authors persuasively advocate for a more holistic approach to understanding bacterial growth regulation, highlighting the complex interplay of genetic factors across the entire genome. Their findings open new avenues for investigating the intricate mechanisms underlying bacterial growth and adaptation, providing fresh insights into bacterial pathogenesis.

Thank you for an excellent summary of our manuscript.

Strengths:

This study exemplifies a holistic approach to unraveling key factors in bacterial pathogenesis. By analyzing a large dataset of whole-genome sequences and employing robust statistical methodologies, the authors provide strong evidence to support their main findings. Which is a leap forward from previous studies focused on a relatively smaller number of strains. Their integration of genome-wide association studies (GWAS) highlights the cumulative, polygenic influences on pneumococcal growth kinetics, challenging the traditional focus on individual loci. This comprehensive strategy not only advances our understanding of bacterial growth regulation but also establishes a foundation for future research into the genetic underpinnings of bacterial pathogenesis and adaptation. The amount of data generated and corresponding approaches to analyze the data are impressive as well as convincing. The figures are convincing and comprehensible too.

Thank you for pointing out the strengths of our manuscript excellently.

Weaknesses:

Despite the strong outcomes of the GWAS approach, this study leaves room for differing interpretations. A key point of contention lies in the title, which initially gives the impression that the research addresses growth kinetics under both in vitro and in vivo conditions. However, the study is limited to in vitro growth kinetics, with the assumption that these findings are equally applicable to in vivo scenarios-a premise that is not universally valid. To more accurately reflect the study's scope and avoid potential misrepresentation, the title should explicitly specify "in vitro" growth kinetics. This clarification would better align the title with the study's actual focus and findings.

Thank you for these suggestions. We have updated the title to include "in vitro" to avoid confusion. The new title now reads, “The capsule and genetic background, rather than specific loci, strongly influence in vitro pneumococcal growth kinetics.” While our study used in vitro data, our goal is to highlight that such in vitro differences in pneumococcal growth may influence in vivo dynamics, as highlighted in several papers referenced in the introduction and discussion.

This study suggests that the entire genetic background significantly influences bacterial growth kinetics. However, to transform these predictions into established facts, extensive experimental validation is necessary. This would involve "bench experiments" focusing on generating and studying mutant variants of serotypes or strains with diverse genomic variations, such as targeted deletions. The growth phenotypes of these mutants should be analyzed, complemented by complementation assays to confirm the specific roles of the deleted regions. These efforts would provide critical empirical evidence to support the findings from the GWAS approach and enhance understanding of the genetic basis of bacterial growth kinetics.

We fully agree with this assessment. As reviewer #1 similarly highlighted, additional genetic manipulation studies would provide further helpful information to assess the impact of altering growth rates on virulence and infection. However, the experimental studies were beyond the scope of this study due to several factors beyond our control. However, we intend to conduct follow-up experimental work to provide additional insights into how the combination of serotypes and genetic background influences pneumococcal growth in vitro and virulence in vivo. Because our current study demonstrates that genetic determinants of pneumococcal growth features are not simply confined to single loci, such experimental validation would require novel wet-lab approaches that consider epistatic interactions.

In the discussion section, the authors state that "the influence of serotype appeared to be higher than the genetic background for the average growth rate" (lines 296-298). Alongside references 13-15, this emphasizes the important role of capsular variability, which is a key determinant of serotypes, in influencing growth kinetics. However, this raises the question: why isn't a specific locus like cps, which is central to capsule biogenesis, considered a strong influencer of growth kinetics in this study?

Thank you for highlighting the point above. Indeed, the capsule biosynthesis (cps) locus is associated with pneumococcal growth kinetics, as seen in the analysis of individual serotypes. However, the cps locus does not come up as a hit in the GWAS because we controlled for the population structure of the pneumococcal strains. The absence of the hits in the cps locus is because serotypes, hence cps loci, tend to be tightly associated with lineages despite occasional capsule switches, which introduce serotypes to different lineages. Therefore, controlling for population structure, which is critical for GWAS analyses, virtually eliminates the detection of potential hits within the cps locus. However, detecting such hits with larger datasets may still be possible. For this reason, we performed a separate analysis of the individual serotypes and lineages shown in Figure 3.

One plausible explanation could be the absence of "elevated signals" for cps in the GWAS analysis. GWAS relies on identifying loci with statistically significant associations to phenotypes. The lack of such signals for cps may indicate that its contribution, while biologically important, does not stand out genome-wide. This might be due to the polygenic nature of growth kinetics, where the overall genetic background exerts a cumulative effect, potentially diluting the apparent influence of individual loci like cps in statistical analyses.

We fully agree with this point. We mentioned in the abstract and discussion that the absence of the signals for specific individual loci within the pneumococcal genome may imply that the growth kinetics are polygenic. We have edited the discussion to emphasise the suggested point.

Reviewer #3 (Public review):

This study provides insights into the growth kinetics of a diverse collection of Streptococcus pneumoniae, identifying capsule and lineage differences. It was not able to identify any specific loci from the genome-wide association studies (GWAS) that were associated with the growth features. It does provide a useful study linking phenotypic data with large-scale genomic population data. The methods for the large part were appropriately written in sufficient detail, and data analysis was performed with rigour. The interpretation of the results was supported by the data, although some additional explanation of the significance of e.g. ancestral state reconstruction would be useful. Efforts were made to make the underlying data fully accessible to the readers although some of the supplementary material could be formatted and explained a bit better.

Thank you for the excellent summary of the manuscript. We have added some text to clarify the significance of some approaches, including ancestral state reconstruction and supplementary material.

Reviewer #1 (Recommendations for the authors):

(1) Since the PCBN was collected pre and post-vaccine introduction, did the authors stratify their analyses other than Figure 7 (disease correlations) to assess how vaccine status may influence growth rates? Is the assertion in Lines 238-239 supported by the in vitro data?

We have done this analysis. Overall, there was no association between vaccine introduction and pneumococcal growth rates. In lines 238-239, we assumed that in vaccinated populations, the host may be more capable of suppressing bacterial replication due to vaccination. However, there was no in vitro data to back this statement. Therefore, we have edited the statement to remove the text regarding vaccination policy.

We considered vaccination status when analysing the data presented in Figure 7. As mentioned in the legend, we only analysed the dataset collected before vaccine introduction to avoid confounding due to vaccination status. To fully assess the impact of vaccination, we would need additional information besides the date of isolation, including vaccine doses and time since vaccination, which was not available for our study.

(2) Similarly, do any of the growth rate metrics correlate with other aspects of the clinical dataset, like the year of isolation or the sex/age of the patient?

We did not include these assessments in the manuscript, as these aspects of the clinical dataset are mostly related to the patient and not necessarily the intrinsic characteristics of the pneumococcus. However, upon revising the manuscript, we compared the growth characteristics against the vaccination period, and we did not find any statistically significant association. The relationship between pneumococcal growth features of the isolates used in the current study and their corresponding clinical manifestations of invasive disease was described in Arends et al. (10.1128/spectrum.00050-22).

(3) When evaluating the impact of serotype on growth rates, did the directionality of some of the described impacts match with those previously reported in other studies?

We were unable to assess the directionality of the serotype’s impact on growth rates. In part, we did not conduct this analysis because our study used different strains from those used in other studies. Such differences in the genetic backgrounds, growth media, and analytical approaches made assessing the consistencies between the studies difficult.

(4) Did the authors expect that a specific growth metric would be more likely to correlate with specific genetic variants? The reader would benefit from a brief discussion of how the metrics (e.g., maximum growth or lag phase duration) are biologically meaningful beyond the overall growth rate.

We indeed expected that specific growth metrics might correlate with certain genetic variants based on their distinct biological roles. The lag phase duration can potentially reflect the ability of the pneumococcus to adapt to environmental conditions, such as nutrient availability or stress, and may be more influenced by regulatory genes involved in sensing and responding to environmental cues (PMID: 30642990, PMID: 22139505). In contrast, maximum growth rate is more likely to be impacted by core metabolic or biosynthetic genes that control the rate of cell division under optimal conditions (PMID: 31053828). Maximum optical density, which reflects the final cell density, might be shaped by factors related to nutrient utilization efficiency, waste tolerance, or quorum sensing. The duration of the stationary phase is related to the switch from lipoteichoic acids to wall teichoic acids, permitting the initiation of the lytic growth phase (PMID: 239401). It is unclear whether this switch is mediated by external triggers or also by intrinsic features of the pneumococcus. Including this type of analysis allows for a more nuanced understanding of how genetic variants contribute to different physiological aspects of microbial growth. The relevance of the lag phase and the stationary phase in relation to the clinical phenotypes of invasive disease (such as pleural empyema and meningitis) of our pneumococcal isolates has been studied and discussed in Arends et al. (PMID: 35678554). The observed associations are summarized in Table 2 of that article. We have added some text in the discussion on the biological relevance of each bacterial growth metric.

(5) For the GWAS analyses, have similar analyses been performed for other S. pneumoniae collections? Are there known "control" loci that the authors could replicate in the current collection to verify the robustness of the approach?

Others have undertaken GWAS analyses of other S. pneumoniae collections elsewhere. Unlike our study, none of the GWAS analyses elsewhere focused on bacterial growth kinetics. Therefore, considering this is the first GWAS study in pneumococcus and bacteria, in general, to focus on growth kinetics, we do not have “control” loci that we could replicate to verify the robustness of the approach. However, we hope that future studies will be able to utilise our findings to compare their approach as more and more similar analyses of in vitro growth data become available.

(6) Is there a statistical method that could predict the sample size necessary to detect the proposed combinatorial or small contributions from various genetic loci to growth rate? This reviewer is not an expert in statistical genetics but would appreciate an indication of the scale required by future studies to identify these regions.

We are unaware of a statistical approach that could predict sample sizes to detect small or combinatorial effect sizes. However, we intend to conduct simulations in future studies to gain insights into the required sample sizes.

(7) WGS and genome assembly metrics should be provided for each sequenced genome especially since only short-read assemblies were performed. If not already deposited, the assemblies should be deposited for data sharing as well.

We have deposited the sequence reads to the European Nucleotide Archive (ENA) and provided the accession numbers, WGS, and assembly metrics in Supplementary Data 1. We have described the tools used to generate the assemblies from the reads.

(8) Please include the specific ethics approval numbers for the sample collection protocol.

Study procedures were approved by the Medical Ethical committees of the participating hospitals, including a waiver for individual informed consent (file number 2020–6644 Radboudumc).

Reviewer #3 (Recommendations for the authors):

Certain aspects of the manuscript could be clarified and extended to improve the manuscript.

(1) Introduction

a) The authors assume knowledge by the reader on Streptococcus pneumoniae, specifically the genetic diversity of lineages and capsules. This diversity is highlighted in the discussion L368 that there are >100 serotypes. The authors should consider backgrounding the number of serotypes and the importance of serotype switching in these bacteria, as well as explaining the diversity of the lineages (GPSC) that are increasingly used as standard nomenclature for Streptococcus pneumonia.

Thank you for bringing this to our attention. We have included a brief description of the GPSC lineages and capsule switching in the introduction.

b) The last paragraph of the introduction is lengthy and gets into the methods and results of the manuscript. These could be edited down.

We have revised the paragraph to remove the methods and results.

(2) Methods

a) The authors should provide details on the QC undertaken and any exclusion criteria of genomes based on the QC. The supplement material has tabs e.g. read and assembly metrics but unclear how determined and impacted the study.

We utilised all the genomes available for this study, which had in vitro phenotypic data available. We excluded no genomes due to poor sequence quality.

Additional information about the genomes is available from previous studies, which are referenced in the methods section.

b) Why did the authors map draft assemblies to the reference genome for the SNP alignment (from which the ML tree was inferred)? Draft genome assemblies usually contain errors so there is potential for false positive SNPs. Further, there is a lack of perbase quality information using the draft genome assemblies. Given the short read data are available - why were the reads not used as input for snippy (which is the standard input for snippy)? This may have impacted the results reliant on the SNP calls.

We mapped a combination of reads and draft assemblies to the reference genome to generate the SNP alignment using Snippy (https://github.com/tseemann/snippy). For the pneumococcal isolates, we mapped the reads, while for the included outgroup, we mapped the assembly as we did not have sequence reads available. We have edited the methods section to clarify this.

c) SNP alignment. the authors explain the decision to not undertake recombination detection later in the discussion. Did the authors mask any phage or repeat regions? And how was the outgroup S. oralis included in the analyses e.g what genome was used?

We included the outgroup genome in the alignment generated by SNIPPY, which involved generating aligned consensus sequences for each isolate after mapping the reads to the pneumococcal ATCC 700669 reference genome (GenBank accession: NC_011900), as described in the methods. We have now included the accession number for the S. oralis genome, which was used as an outgroup in our phylogenetic analysis. Phages are not typically common in pneumococcal genomes compared to other species. Similarly, although repeats are present in the pneumococcal genome, the consensus in the field is that these do not particularly bias the pneumococcal phylogeny. Therefore, the consensus in the field has been not to explicitly mask these regions as done for highly clonal bacterial pathogens, such as Mycobacterium tuberculosis. Overall, our approach to building the phylogenetic tree is robust compared to alternative methods (PMID:

29774245).

d) Should the presence/absence of unitigs that were used as the input for the GWAS be included as a supp dataset?

We have now provided the presence/absence matrix for the unitigs used in the analysis as a supplementary dataset available at GitHub(https://github.com/ChrispinChaguza/SpnGrowthKinetics). We have revised the methods section to include a section on data availability.

e) For the annotation of unitigs, the authors used their bespoke script with features from complete public genomes. Please provide accession/ identifying information of the complete genomes (not only the ATCC 700669) reference in the methods. Also, why did the authors choose not to annotate with annotate_hits_pyseer from pyseer?

We annotated the hits using our bespoke script because we understood our approach better and could control the information generated from the script. Annotating with “annotate_hits_pyseer” from pyseer would produce similar results to both approaches, as they compared the unitigs to annotated reference genomes.

(3) Results

a) The authors could consider providing an overview of the diversity (e.g. lineages and capsules) in the study and contextualising it in the broader context of Streptococcus pneumoniae population genomics. This would help readers who are less familiar with this pathogen to understand the diversity included in this study.

We included this information in the first paragraph of the results section. Considering that population-level analyses based on this dataset have already been published, we have referenced the corresponding papers to provide additional information to readers.

b) Did the timespan of the study pre and post-PCV7 introduction need to be briefly touched on in the results? For example, did the serotypes and lineages vary over the two collection periods and does this need to be considered in the interpretation of the results at all?

The prevalence of serotypes and lineages varied over time, partly due to the introduction of vaccines and random temporal fluctuations in the distribution of strains. We did not explicitly adjust for time, as this is not likely to influence the intrinsic biology of the strains. However, we adjusted for the population structure of the strains, whose changes would most likely affect the distribution of strains in the population. For other analyses, including that in Figure 7, we considered the vaccination status by restricting the analysis to the isolates collected before vaccine introduction.

c) Figures. Some of the figures had very small text (especially Figure 1) that was difficult to read and Figure 2 and Figure 4 were mentioned once, while several paragraphs of results were used to discuss Figure 3. Is Figure 1 required as a main figure? Could Figure 3 be split? e.g. one with the chord diagram, one with panels b-e, and one with panels jq? Figure 4 - the ancestral state reconstruction analyses could be expanded upon in the results.

We have increased the text in some figures where possible. However, for figures that show more information, smaller text is more suitable.

Figure 1 is essential to the manuscript as it provides a visual overview of the approach used in this study. Without this figure, it may be difficult for some readers, especially those unfamiliar with bacterial genomic analyses, to understand our study approach and how we estimated the pneumococcal growth parameters used for the GWAS.

For Figure 4, we prefer to keep it as it is, to have the information in one place, as splitting it will mean including some of the panels in the supplementary material, considering that we already have seven figures in the manuscript.

We have added additional text to the results regarding the ancestral reconstruction analyses. We included them mainly to demonstrate the correlation between the pneumococcal growth rates and the phylogeny.

(4) Discussion

a) Why was 15 hours for culture undertaken and not 24? The authors discuss the impact that this may have had on their results.

The 15-hour incubation period was deliberately chosen, as the growth curves indicate that most isolates had reached the stationary phase by that time. Extending the culture duration would likely not have yielded additional meaningful data. As is well established, Streptococcus pneumoniae undergoes autolysis upon reaching a certain cell density, which could distort growth measurements and complicate interpretation if incubation were prolonged. For clarification, we have changed the sentences related to this topic in the Discussion.

b) Some paragraphs in the discussion were very long e.g. L347-381. The authors could consider breaking long paragraphs down into shorter ones to improve the readability of the manuscript.

We agree with this assessment. We initially wanted to include all the information on the study’s limitations in the same paragraph. However, as suggested, we have now split the highlighted paragraph into two shorter paragraphs.

(5) Supplementary Data

a) Providing information in each tab of each supp data file would be useful. For example - including a table header that explained what was in each sheet rather than relying on the tab names. Formatting for some of the underlying supplementary data could be improved e.g. in supplementary data 2 no explanation is given to interpret the data included in these files.

Thank you for the suggestions. For clarity, we have included a header in each tab of the spreadsheet that describes what is included in each dataset. We have also removed the previous Supplementary Data 2. We realised that the information presented in this spreadsheet was redundant, as it was already available in Supplementary Data 1.

Version 2

Abstract

Abstract

Bacterial growth characteristics intrinsic to each strain can impact and influence gene expression, antibiotic susceptibility, and disease pathogenesis. However, little is known about specific genomic variations that influence these bacterial growth features. Here, we investigate the impact of Streptococcus pneumoniae genetics on its in vitro growth features to shed light on genes that may be important targets in the development of vaccines and therapeutics. We use statistical models to estimate growth features and demonstrate that they varied significantly across capsular serotypes and lineages, were strongly correlated with phylogeny, and showed high heritability, highlighting a strong genetic basis. Despite this, genome-wide association studies revealed no specific genomic loci statistically associated with the growth features independently of the genetic background, including those in the locus responsible for capsular polysaccharide synthesis. Our findings suggest that the serotype and lineage, as well as a combination of genomic loci, influence intrinsic pneumococcal growth kinetics, which may have implications for pneumococcal disease pathogenesis.

Introduction

Once termed “the captain of the men of death” by Sir William Osler in the 20th century, Streptococcus pneumoniae, or the pneumococcus, remains a significant cause of life-threatening invasive diseases, such as sepsis, pneumonia, and meningitis1, despite the widespread implementation of life-saving serotype-specific polysaccharide pneumococcal conjugate vaccines (PCVs)2. The high capsular diversity, comprising ∼100 serotypes with varying potential for colonising the nasopharynx and causing invasive diseases3,4 and mortality5–7, demonstrates the remarkable genetic diversity of the pneumococcus. Such high genetic diversity reflects the plasticity of the pneumococcal genome, which readily exchanges genetic material with other strains through recombination and horizontal gene transfer8. Capsule switching, resulting from the swapping of the capsule biosynthesis (cps) loci, which encode the extracellular polysaccharide capsule that determines the capsular serotype, exemplifies genetic exchange between pneumococcal strains9. While serotypes typically exist in specific lineages10, such as those defined by the pneumococcal global sequence cluster (GPSC) nomenclature11, some serotypes are expressed across distinct lineages, while some lineages express multiple serotypes, reflecting capsule switching12. However, the impact of individual capsular serotypes, genetic background, and their combinations on the bacterial phenotypes, especially those associated with disease pathogenesis and clinical manifestations, remains less understood.

The pneumococcus possesses an arsenal of virulence factors. The extracellular polysaccharide capsule is a primary virulence factor crucial for pneumococcal transmissibility, survival, and pathogenicity in the human host13–15. Recent studies on the pneumococcus and other bacterial pathogens suggest that variability in growth kinetics affects disease pathogenesis, clinical manifestation, and antibiotic therapy. For example, differential pneumococcal growth features appear to correlate with fitness for colonisation and invasive disease,16 clinical manifestation16–18 and response to environmental conditions18–20. Studies of other bacterial pathogens suggest that growth rate influences gene and protein expression21,22, adhesion23, bacterial competition24, plasmid replication25, bacteriophage dynamics26, and antibiotic accumulation and efficacy27–29.

Similarly, experimental and mathematical modelling studies have demonstrated that bacterial growth correlates with bacterial load and host-mediated cell lysis, impacting dispersal to different host cells and tissues during systemic infection30. Previous studies of pneumococcus and other bacteria have suggested the influence of environmental factors19,20 and bacterial-specific factors, including capsular serotype18,31 and the presence of plasmids32 on bacterial growth phenotypes. These growth phenotypes may consequently impact the pathogenesis of invasive pneumococcal disease16–18. However, the specific pneumococcal genomic variations affecting growth kinetics remain less understood.

Studying bacterial growth kinetics sheds light on the influence of genes, genomic variation, and growth media conditions on bacterial fitness33,34. However, studying bacterial growth kinetics in vivo remains intrinsically challenging due to the complexity of the conditions in the hosts, although innovative approaches are being developed35,36. On the other hand, in vitro methods provide a valuable alternative for assessing the impact of specific bacterial and environmental factors on bacterial fitness. Although these in vitro methods have yielded remarkable insights into bacterial fitness, they have primarily provided strain-specific rather than population-level insights due to the analysis of a small number of isolates from a limited number of genetic backgrounds.

Here, we studied the impact of pneumococcal genetics on in vitro growth kinetics to understand the influence of bacterial genetic variation on pneumococcal pathogenesis. Specifically, we performed high-throughput quantification of the in vitro growth kinetics and whole genome sequencing of 348 invasive pneumococcal isolates from the Netherlands to determine the impact of genomic variation on growth kinetics. We hypothesised that the capsular serotype, genetic background, and specific individual genomic loci independent of the genetic background influence in vitro pneumococcal growth kinetics features or parameters derived from the growth curves. We tested this hypothesis through a series of statistical and comparative genomics analyses, including a bacterial genome-wide association study (GWAS) approach. Our study provides insights into the influence of capsular serotype, genetic background or lineage, and individual genomic loci on pneumococcal growth kinetics – a fundamental trait that potentially influences nasopharyngeal colonisation, invasive disease pathogenesis, and clinical manifestations.

Results

Population structure of a diverse collection of Dutch invasive pneumococcal isolates and derivation of their growth kinetics

We aimed to understand the impact of the capsular serotype, genetic background, and specific genomic loci, independent of the clonal population structure, on the pneumococcal growth features (Fig. 1 and Supplementary Data 1). Therefore, we investigated the in vitro growth kinetics of 348 pneumococcal isolates collected from patients with invasive pneumococcal diseases via the Pneumococcal Bacteraemia Collection Nijmegen (PBCN) cohort study at two hospitals in Nijmegen, the Netherlands before (2000-2006) and after (2007-2011) the introduction of the seven-valent pneumococcal conjugate vaccine (PCV) [PCV7] into the Dutch paediatric immunisation programme37. Our isolates represent 34 capsular serotypes, 40 pneumococcal lineages, as defined by the global pneumococcal sequence cluster (GPSC) nomenclature11, and 99 sequence types (STs) specified by the pneumococcal multilocus sequence typing (MLST) scheme38. An interactive and annotated maximum likelihood phylogeny of the isolates used in this study is available online at the Microreact platform for open data visualisation and sharing for genomic epidemiology (https://microreact.org/project/wxmdblfgprepwugvbbtaq7). We estimated a Simpson diversity index of 0.932 (95% confidence interval [CI]: 0.924–0.940), 0.937 (95% CI: 0.928–0.945), and 0.958 (95% CI: 0.949–0.966) for the serotypes, lineages, and STs, respectively. The high richness and diversity of pneumococcal serotypes and lineages highlighted a substantial population-level diversity of the study isolates in the Netherlands, consistent with earlier studies of the same population39,40.

Schematic overview of high-throughput quantification of in vitro pneumococcal growth kinetics and the genome-wide discovery of individual genomic variation influencing in vitro pneumococcal growth.

Population genomic approaches using genome-wide association study (GWAS) offer a robust way to identify genomic variation influencing the growth dynamics of the pneumococcus. First, we collected 348 pneumococcal blood culture isolates that were obtained from 348 patients admitted to the hospital with invasive pneumococcal disease. The isolates were inoculated in a liquid medium in a highly standardised manner followed by measurement of the growth kinetics after every ten minutes for slightly over 15 hours using a microplate reader17. We then fitted a logistic function (see methods) to measure growth features, namely, the average growth rate (r), maximum growth density (Hmax) defined as Lmax, lag phase duration, and maximum growth change (ΔH), defined as the difference between the maximum and minimum density (Lmax-Lmin). Additionally, genomic DNA was extracted from the cultured isolates for whole genome sequencing using the Illumina HiSeq 4000 sequencing platform. The generated sequencing data was used to reconstruct a maximum-likelihood phylogenetic tree of the isolates for downstream analyses, for example, the assessment of phylogenetic signals for the estimated growth features. We also performed de novo assembly of the sequencing reads to generate contigs, which were used to map against the ATCC 700669 pneumococcal reference genome (GenBank accession: NC_011900) to identify consensus single nucleotide polymorphisms (SNPs). We also generated presence, and absence patterns for unitigs or variable length k-mer sequences based on the draft assembled contigs. The SNPs were used to generate the phylogeny of the isolates while the unitigs were used in a GWAS to identify variants associated with the estimated pneumococcal growth features. The images were created with (BioRender).

Capsular serotype and genetic background significantly impact pneumococcal growth kinetics

To explore the impact of capsular serotype and genetic background or lineage on the pneumococcal growth features estimated from a fitted growth curve, namely lag-phase duration, maximum growth density (Hmax), maximum growth change (ΔH), and average growth rate (r) (Fig. 1, 2, and Supplementary Data 1), we assessed the relationship between these growth features and serotypes and lineages associated with the study isolates (Fig. 3a). In general, we found significant variability of the growth features across serotypes: average growth rate, maximum growth density, maximum growth change, and lag phase duration (P<0.0001; Kruskal-Wallis test) consistent with previous studies18,19 (Fig. 3b-e). We observed similar patterns across different pneumococcal lineages (Fig. 3f-i and Supplementary Data 2).

Growth curves for selected serotypes show the in vitro pneumococcal growth dynamics and estimated growth features inferred by mathematical modelling.

The growth curves above show the optical density at 620 nm monitored over 15 hours for an example dataset of 20 pneumococcal isolates representing different serotypes and sequence types (STs) based on the pneumococcal MLST scheme. We used the nonlinear optimisation algorithm to identify a combination of parameters with the best fit to a logistic curve based on the lowest residual sum of squares for the predicted value based on the model relative to the observed growth data (see methods). We fitted separate curves based on a subset of points for each replicate up to the end of the stationary phase or end of the experiment, whichever came first. We did this to get a better fit for the logistic curve during the growth phase, as it does not allow for negative growth, which happened after the stationary phase. We fitted the curves to infer the growth features for each of the six replicates. The averaged estimates for each replicate were then used for the downstream analyses. Additional plots showing fitted curves for all the isolates are shown in Supplementary Fig. 1.

The in vitro pneumococcal growth features show variable patterns dependent on the capsular serotype and genetic background or lineage.

(a) Chord diagram showing the association between pneumococcal capsular serotypes and lineages for the Dutch isolates included in this study. The serotypes were determined using in silico approaches based on the sequencing data using SeroBA91, while the lineages were defined using the PopPUNK framework92 based on the global pneumococcal sequencing cluster (GPSC) lineage nomenclature11. The thickness of the connecting lines between serotypes and lineages in the chord diagram represents the percentage of isolates belonging to each serotype and lineage. Due to capsule switching, some serotypes were associated with multiple lineages, and similarly, some lineages were associated with multiple serotypes. The violin plot shows the distribution of the estimated average growth rate across different serotypes. The distribution of the estimated growth features varied across serotypes with at least three isolates, as shown by the violin plots for the (b) average growth rate, (c) maximum growth change, (d) maximum growth change across serotypes, (e) lag phase duration growth features. Similarly, the distribution of the estimated growth features varied across lineages with at least three isolates, as shown by the violin plots for the (f) average growth rate, (g) maximum growth change, (h) maximum growth density, and (i) lag phase duration growth features. A comparison of the growth features between two serotypes belonging to the same lineages with at least three isolates is shown for the (j) growth rate, (k) maximum growth, (l) maximum growth change, and (m) the lag phase duration features. A comparison of the growth features between two lineages expressing the same serotype containing at least three isolates are shown for the (n) growth rates, (o) maximum growth, (p) maximum growth change, and (q) the lag phase features. Additional information regarding the isolates and their estimated growth features or parameters are available in Supplementary Data 2.

Recombination and horizontal gene transfer may lead to the replacement of part or the entire capsule biosynthesis locus, resulting in the production of a polysaccharide capsule distinct from the original one, a phenomenon called capsule-switching41. Consequently, some lineages are associated with multiple serotypes, and similarly, a single serotype can be expressed in distinct genetic backgrounds12. Examples of lineages containing multiple serotypes included GPSC3, GPSC4, and GPSC7, and likewise, serotypes, including 3, 19A, and 19F, were expressed in multiple genetic backgrounds, which also express other serotypes (Fig. 3a). Overall, the average growth rate for serotypes expressed in different genetic backgrounds broadly showed similar patterns. For example, lineages expressing serotype 14 had consistently higher average growth rates. In contrast, lineages expressing serotype 3 showed lower rates, consistent with the notion that capsule production interferes with pneumococcal growth18 (Fig. 3b). Therefore, we investigated the impact of the capsular serotype and genetic background on the estimated pneumococcal growth features.

First, we identified lineages that expressed multiple serotypes and serotypes found across multiple lineages. To ensure a robust statistical comparison, we only focused on serotype-lineage combinations with at least three sequenced isolates in our study. We identified two lineages, GPSC3 and GPSC7, with sufficient sequenced isolates for comparison, which expressed multiple serotypes (Fig. 3j). We observed higher growth rates for serotype 33F than serotype 8 (P=0.0007; Kruskal-Wallis test).

In contrast, we found no statistically significant differences in growth rates between serotypes 23A and 23F in lineage GPSC7 (P=0.2568; Kruskal-Wallis test). We speculated that the absence of statistically significant differences between serotypes 23A and 23F reflected their presence in the same serogroup; therefore, they had similar capsule properties. Interestingly, we observed similar patterns for the maximum growth density, maximum growth change, and lag phase duration features between serotypes on the GPSC3 and GPSC7 lineages (Fig. 3k-m and Supplementary Data 2). Next, we investigated the growth rates of the same serotypes expressed in distinct genetic backgrounds. Although some serotypes exhibited similar growth rates across distinct genetic backgrounds, such as serotype 3 (P=0.4599; Kruskal-Wallis test), others showed different growth rates across different lineages, as seen with serotypes 14 (P=0.0006; Kruskal-Wallis test) and 19F (P=0.0105; Kruskal-Wallis test). We noted similar patterns for the other growth features: maximum growth density, maximum growth change, and lag phase duration (Fig. 3n-q).

These findings demonstrate that the capsular serotype and genetic background significantly impact pneumococcal growth kinetics. The resultant growth features likely depend on the interaction between these factors, whereby some combinations may result in more pronounced growth effects than others.

Pneumococcal growth kinetics show heritability and correlation with the phylogeny of the isolates

Based on the findings that both capsular serotype and genetic background are associated with growth kinetics, we hypothesised that the estimated pneumococcal growth features exhibit a high correlation between the pneumococcal growth features and the phylogeny, i.e., phylogenetic signal. To assess this, we first performed ancestral state reconstruction to map the derived pneumococcal growth features or parameters across the whole genome phylogeny of the isolates. By reconstructing the estimated growth features on the internal nodes of the phylogeny, we found high but variable associations between the growth features and different genetic backgrounds (Fig. 4, Supplementary Fig. 2 and 3, and Supplementary Data 2). Broadly, genetically similar isolates from the same phylogenetic lineages or genetic backgrounds exhibited similar growth kinetics, varying in degree depending on the specific growth feature. Altogether, these findings suggest a strong phylogenetic signal in the growth features.

Ancestral reconstruction reveals a strong correlation between the estimated in vitro pneumococcal growth features and the whole-genome phylogeny of the isolates.

The evolution of the normalised estimated continuous growth features, namely average growth rate, maximum growth change, maximum growth density, and lag phase duration, were assessed by estimating the states of the internal nodes of the maximum likelihood phylogenetic tree of the isolates using a maximum likelihood ancestral reconstruction approach98. The length of the scale bar shows the number of nucleotide substitutions per site, while the colours represent the lowest and highest unnormalised growth features. The length of the scale bar shows the number of nucleotide substitutions per site, while the colours represent the lowest and highest unnormalised growth features. A phylogenetic tree showing bootstrap support values on the internal nodes of the tree is provided in Supplementary Data 3.

Next, we calculated the phylogenetic signal of each pneumococcal growth feature using Pagel’s λ statistic38, a metric typically ranging from 0 to 1 that evaluates the correlation between given traits and the phylogeny, with higher values indicating a stronger phylogenetic signal. The amount of the phylogenetic signals based on the normal transformed growth features was highest for the average growth rate (Pagel’s λ=0.72, 95% CI: 0.61 to 0.81; P<2.08×10−64), maximum growth change (Pagel’s λ=0.60, 95% CI: 0.40 to 0.76; P<1.53×10−16), and maximum growth density (Pagel’s λ=0.73, 95% CI: 0.57 to 0.84; P<4.18×10−26), while the lag phase duration showed a lower phylogenetic signal (Pagel’s λ=0.30, 95% CI: 0.17 to 0.47; P<5.42×10−16) (Fig. 5a). These findings suggest that there is a heterogeneous correlation of the growth features with the phylogeny.

The estimated in vitro pneumococcal growth features show high phylogenetic signals and narrow-sense heritability.

(a) Bar plot showing the phylogenetic signal of the growth features using Pagel’s λ statistic42. Pagel’s λ measures the correlated evolution of the traits with the phylogenetic tree of the isolates. Pagel’s λ estimates close to zero indicate no phylogenetic signal, i.e., the growth features were independent of the phylogeny, while values close to one show a strong phylogenetic signal. The average growth rate, maximum growth change, and maximum growth density showed the highest phylogenetic signal, while the lag phase duration feature showed an intermediate signal. In contrast, the maximum growth rate defined by Arends et al.17 had the lowest phylogenetic signal, i.e., evolved more independently of the phylogeny. (b) Bar plot showing the narrow-sense heritability (h2) of the pneumococcal growth features based on unitig sequences. The narrow-sense heritability corresponds to the amount of variability in the traits, i.e., growth features, explained by the pneumococcal genetic variation. The heritability is the proportion of phenotypic variation explained by the bacterial genetics (PVE) parameter using the linear mixed effects model implemented in GEMMA102. A heritability value close to zero implies low heritability or minimal contribution of genetics to the variability of the traits. In contrast, values close to one suggest a strong influence of genetics on growth features. There was a substantial impact of pneumococcal genetics on all the growth features, with the highest influence seen for the average growth rate, maximum growth change, and maximum growth density, with a slightly lower contribution for the lag phase duration and the maximum growth rate measured by Arends et al.17. All the error bars represent the 95% confidence intervals. Additional information is provided in Supplementary Data 3.

To quantify the variability in pneumococcal growth kinetics explained by genotype, we calculated the narrow-sense heritability (h2) for the four normal-transformed estimated growth features and the maximum growth rate from Arends et al.17. We found strong, though not equal, evidence for a genetic basis for all the estimated growth features: average growth rate (h2=0.76, 95% CI: 0.68 to 0.85), maximum growth change (h2=0.80, 95% CI: 0.71 to 0.89), maximum growth density (h2=0.88, 95% CI: 0.83 to 0.94), and lag phase duration (h2=0.29, 95% CI: 0.14 to 0.43) (Fig. 5b, Supplementary Data 2). To further understand the impact of the capsular serotype and genetic background, we fitted a linear mixed effects model for each growth parameter with serotype and lineage (defined by the GPSC nomenclature11) as random effects to estimate the amount of variance of the growth features explained by the serotype (h2 serotype) and lineage or GPSC (h2GPSC) calculated as the ratio of the variance of the serotype and lineage components relative to the total variance, including the residual or environmental effects. We calculated h2serotype =0.47 and h2GPSC =0.11 for the average growth rate, h2serotype =0.13 and h2GPSC =0.28 for the maximum growth change, h2serotype=0.29 and h2GPSC=0.46 for the maximum growth, h2serotype =0.08 and h2GPSC =0.39 for the lag phase duration feature.

Altogether, these findings demonstrate a substantial impact of serotype and lineage on the variation in pneumococcal growth features, whereby most serotypes tend to have similar characteristics regardless of the genetic background. Similarly, some genetic backgrounds tend to exhibit similar phenotypic patterns regardless of the serotypes, but the serotype appears to have a more significant influence on the growth characteristics.

No individual genomic loci are associated with pneumococcal growth features

After finding moderate to high narrow-sense heritability of the growth features, we then undertook a GWAS analysis to identify specific genomic loci associated with these features, independent of the strains’ genetic background using a linear mixed-effects model. We performed separate GWAS analyses of the maximum growth change, maximum growth density, maximum growth change, and lag phase duration phenotypes derived from the growth curve. We conducted the GWAS using variable-length k-mer sequences or unitigs to capture genetic variation in the pneumococcal genomes at different levels, including SNPs, multi-allelic sites, and insertions and deletions43,44. We found no individual unitigs associated with the derived growth feature phenotypes (Fig. 6a). Furthermore, repeating the GWAS analyses adjusting for the serotype revealed no statistically significant hits associated with the derived growth phenotypes, suggesting that the variants in the pneumococcal strains are correlated with the growth kinetics independent of the serotype (Fig. 6b). Overall, the observed and expected statistical significance values in the QQ-plots for our study revealed no apparent issues with the population structure of the isolates, which is a significant confounder in similar studies45 (Fig. 6c, d). While the capsule is biologically relevant for pneumococcal growth (Fig. 3b-q), the absence of individual loci even within the capsule biosynthesis locus with elevated signals associated with the growth features may be a mere consequence of adjusting for the population structure as pneumococcal serotypes are typically associated with specific lineages (Fig. 3a).

Pneumococcal genetic variation associated with the in vitro growth features or parameters of the isolates.

Manhattan plot showing the statistical significance of the hits from the GWAS analysis based on unitigs (a) without adjusting for the serotype and (b) after adjusting for the serotype. The statistical significance (P-value) were inferred based on the likelihood ratio test using Pyseer100 mapped against the ATCC 700669 pneumococcal reference genome (GenBank accession: NC_011900107) to identify their location in the pneumococcal genome. The dots in the Manhattan plots with different colours represent GWAS for different growth features, as shown in the key. The red dotted line designates the genome-wide threshold for considering a variant as significantly associated, statistically, with the growth features. The blue dotted line represents the threshold for classifying variants as suggestively associated with the growth features. (c) Quantile–quantile plots showing the relationship between the observed statistical significance and the expected statistical significance for the GWAS of growth features not adjusting for the serotype. (d) Quantile–quantile (QQ) plots showing the relationship between the observed statistical significance and the expected statistical significance for the GWAS of growth features after adjusting for the capsular serotype.

Overall, these findings suggest that the contribution of individual genomic loci in influencing pneumococcal growth kinetics does not stand out, potentially reflecting the polygenic nature of the phenotype and cumulative effect of the genetic background or lineage.

Average in vitro growth rates are not strongly correlated with a serotype’s invasiveness or its prevalence in carriage or disease

Previous studies have elucidated that bacterial growth rate may influence colonisation and disease20, including systemic dissemination during infection30 and disease manifestations17,31.

Pneumococcal strains that are prolific growers may exhibit better survival and replication in systemic tissues than those that are slow growers. We used isolates collected before the introduction of the PCV7 vaccine in the Netherlands46 to assess the association between in vitro growth rate of the isolates and their population-level characteristics: prevalence in invasive disease and nasopharyngeal carriage and invasiveness, i.e., the odds of being isolated in invasive disease compared to the carriage state. Our first hypothesis was that the average in vitro growth rate for each pneumococcal serotype is positively associated with the prevalence of that serotype in invasive diseases. Our analysis revealed no evidence of an association between the prevalence of serotypes in invasive disease and the in vitro growth rate (Spearman correlation: –0.36, P=0.098) (Fig. 7a, Supplementary Data 3). Second, we hypothesized that the average in vitro growth rate of the serotypes is positively associated with the prevalence of the serotypes in the nasopharyngeal carriage, as fitter strains would grow faster than less fit strains. Similarly, we found no association between the prevalence of serotypes in carriage and the in vitro growth rate (Spearman correlation=-0.41, P=0.12) (Fig. 7b, Supplementary Data 3). We repeated the analysis after excluding serotype 3 and found no evidence of an association between growth rate and carriage prevalence (Spearman correlation=-0.28, P=0.314), which suggests that serotype 3 alone was driving the previously observed association. Finally, we tested the hypothesis that the growth rate is positively associated with the invasiveness of the serotypes. We found no association between the invasiveness and the in vitro growth rates of the serotypes (Spearman correlation: 0.144, P=0.594) (Fig. 7c and Supplementary Data 3). These findings suggest that growth rates alone may not be sufficient to explain the prevalence and invasiveness of pneumococcal serotypes, indicating the influence of other pneumococcal factors, microbiome (e.g., viral infection), and host-related factors in determining pneumococcal disease susceptibility.

Association between average in vitro pneumococcal growth rate with the prevalence of serotypes in invasive disease, nasopharyngeal carriage, and invasiveness.

(a) Scatter plot showing the relationship between the prevalence of serotypes among invasive disease isolates collected before the introduction of PCV7 in the Netherlands and the average in vitro growth rate (r) of the isolates inferred as shown in Fig. 1. A Spearman correlation test showed no evidence of an association between the growth rate and prevalence of serotypes among the invasive disease isolates. (b) Scatter plot showing the relationship between the inferred prevalence of serotypes in the nasopharyngeal carriage before the introduction of PCV7 in the Netherlands and the inferred average in vitro growth rate. Due to the unavailability of the nasopharyngeal carriage data, we imputed the prevalence of the serotypes in carriage by dividing the prevalence in invasive disease by the average invasiveness of each serotype. The invasiveness index was calculated based on the meta-analysis of data collected before PCV7 introduction across multiple countries (see methods). A Spearman correlation test of the growth rate and prevalence of serotypes in carriage suggested a negative association between the variables. The association remained even after excluding serotype 3, which seemed like an outlier. (c) The relationship between the invasiveness and the average growth rates of each serotype from the pre-vaccination data. A Spearman correlation test of the growth rates and invasiveness suggested no association between the variables. All the error bars represent the 95% confidence intervals. Additional information is provided in Supplementary Data 3.

Discussion

Bacterial growth kinetics impact several traits, including gene and protein expression21,22, antibiotic accumulation and efficacy27–29, cell surface adhesion23, intra– and inter-specific competition24, response to environmental conditions18–20, plasmid replication25, bacteriophage dynamics26, colonisation and disease pathogenesis16, including clinical manifestation16–18. Whilst previous studies have attempted to understand bacterial, host, and environmental factors influencing bacterial growth, a major drawback has been a narrow focus on a small subset of strains47. Therefore, the insights gained from these studies have been mostly limited to specific strains, as they ignored potential variability across different genetic backgrounds. Such population-level studies are critical for genetically diverse bacterial pathogens with highly structured populations11, substantial antigenic diversity48, and recombination rates, such as the pneumococcus8,10,41. Here, we investigate the association between pneumococcal genetics and growth kinetics using a combination of high-throughput quantification of in vitro pneumococcal growth, whole-genome sequencing, and population genomic analyses employing a bacterial GWAS approach. Notably, our findings reveal a strong influence of the capsular serotype compared to the genetic background, and the absence of specific individual genomic loci modulating pneumococcal growth kinetics, particularly the in vitro growth rate. These findings may partially explain why some serotypes and lineages colonise and cause invasive diseases more effectively than others and may have implications for developing measures to prevent and control pneumococcal infections.

Our findings suggest that the capsular serotype and genetic background influence the growth kinetics of pneumococci. The findings are generally consistent with results from experimental studies by others17–19, which used capsule-switched strains to exclude the impact of the genetic background and demonstrated an alteration in growth dynamics due to the capsule. However, there may be potential differences in the growth media and conditions. Specifically, encapsulated strains exhibited higher growth rates than unencapsulated strains in nutrient-rich conditions18.

These findings suggest that the capsule is crucial for pneumococcal growth, but its effect depends on nutrient availability – promoting growth when nutrients are abundant and interfering with growth when nutrients are limited, respectively18,31. A recent study by Tóthpál et al.19 suggested that the serotype partially explains pneumococcal growth and suggested the genetic backgrounds, which vary in their ability to colonise the nasopharynx and cause invasive diseases4,49–51, may also play a role. However, the authors did not fully quantify the impact of the genetic background as the isolates were not sequenced19. Our study confirms the strong influence of both the pneumococcal capsular serotype and genetic background or lineage on the growth kinetics of the isolates. The influence of the serotype appeared to be greater than that of the genetic background on the average growth rate, while the other growth features seemed to show an opposite trend. These findings suggest that the respective influence of the capsular serotype and lineage varies by specific derived pneumococcal growth features.

Interestingly, although our findings revealed that the capsular serotype and genetic background influence pneumococcal growth kinetics, statistical analysis showed that the phenotypic variability attributed to the capsular serotype was greater than that attributed to the genetic background. Specifically, the capsular serotype exhibits a more pronounced growth rate, maximum growth change, and maximum growth density than the genetic background. In contrast, the genetic background had a greater effect on the lag phase duration than the serotype, a growth phase suggested to be responsible for bacterial adaptation to new environments52. We acknowledge that our comparison of growth features in the same genetic background may also indicate a significant amount of genetic variation for some of the lineages, some of which could be responsible for changes in growth kinetics. These findings illustrate the combined, albeit variable, importance of the capsular serotype and genetic background on pneumococcal growth kinetics, which may partially explain the pathogenesis and prevalence of the pneumococcus in nasopharyngeal carriage and invasive disease.

Previous studies have revealed individual bacterial genomic loci that impact other phenotypes, including pathogenicity53–61, virulence62–65, host, niche, and environmental adaptation56,58,66, conjugation67, and antimicrobial resistance68–75, independent of the genetic background.

Although our study focused primarily on the pneumococcal growth rate parameter, the other growth parameters are also biologically relevant. For example, the lag phase duration potentially reflects the ability to adapt to environmental conditions, such as nutrient availability or stress, and may be more influenced by regulatory genes involved in sensing and responding to environmental cues52,76. On the other hand, the maximum growth rate may be impacted by genes controlling the rate of cell division under optimal conditions77 while the maximum growth density, which reflects the final cell density, might be shaped by factors related to nutrient utilization efficiency, waste tolerance, or quorum sensing. The duration of the stationary phase may reflect the initiation of the lytic growth phase, possibly due to the effects of an autolysin inhibitor78. Our GWAS analysis, undertaken to gain a more nuanced understanding of how genetic variants contribute to different physiological aspects of microbial growth, revealed no elevated signals in specific genomic loci associated with pneumococcal growth features. This suggests that individual variants may not significantly influence pneumococcal growth kinetics independently of the genetic background and serotype. Previous studies have shown that mutations in other capsule biosynthesis genes (wchA or cpsE) encoding a protein that links activated glucose phosphates to lipid carriers48 also influence the growth dynamics of the pneumococcus47. However, our results contradict our initial speculation that genetic variation in the capsule polysaccharide synthesis genes would be strongly associated with pneumococcal growth features. These findings suggest that the contribution of specific individual pneumococcal genomic loci to growth kinetics does not stand out, hinting at the potential polygenic nature of growth kinetics and an overall impact of the genetic background, which exerts a cumulative effect, diluting the influence of individual loci. Uncovering such effects will require follow-up studies with larger sample sizes.

Compared to the findings of a previous study by Hathaway et al.18, we found no robust evidence for an association between the in vitro growth rates and the population-level prevalence of the serotypes in nasopharyngeal carriage. However, this may have occurred as a result of a limitation of our study, since we studied a collection of randomly sampled invasive isolates, and the growth experiments were conducted under conditions similar to those of blood. In contrast, Hathaway et al.18 used a non-random selection of both carriage and disease isolates. Isolates likely have adaptations to the environmental conditions in which they were isolated; therefore, in vitro growth characteristics for any given isolate may be different under experimental conditions simulating different niches. In addition, the in vitro growth environment cannot fully represent in vivo conditions and fails to consider the impact of selective pressures arising from serotype-specific responses (antibodies) or non-specific responses (phagocytosis) as well as potential interactions with other pathogens. The lack of association between the invasiveness and prevalence of serotypes in invasive disease and carriage, as well as the growth rates, could reflect that the conditions in the in vitro growth environment do not fully represent in vivo conditions.

Collectively, these findings suggest that these and other factors may substantially influence the complex pathogenesis of pneumococcal disease in vivo.

The strength of this study lies in the analysis of an extensive collection of well-phenotyped natural isolates, combined with phenotyping and the estimation of growth features using statistical models and statistical genetics to determine the association between the growth features and genetic variation across the entire genome. As previously noted by Tothpal et al.19, the culturing history of the isolates may impact the growth kinetics. In our study, we consistently cultured all isolates after isolation from patients using standardised culturing procedures (single assay, identical initial concentration, environmental conditions (temperature), and equipment to measure optical density, and repeated measurements to spot technical variability17). Compared to the present study, future studies would benefit from larger sample sizes and may yield more results, as we only explained a portion of the heritability of the growth features. We also acknowledge that the methods used to estimate heritability may also sometimes perform poorly on certain bacterial species; therefore, more robust methods are needed to validate our findings further79. We conducted our in vitro culture experiments for 15 hours, as the growth curves indicate that most isolates had reached the stationary phase by that time. Increasing the culturing time from 15 to >24 hours, as done elsewhere14,15, might be considered for follow-up studies. However Streptococcus pneumoniae undergoes autolysis upon reaching a specific cell density, which could distort growth measurements and complicate interpretation if incubation were prolonged. Furthermore, since the pneumococcus invades different niches and systemic tissues under varying conditions, including oxygen concentration and temperature16,18–20, future studies should investigate the effect of these environmental conditions on pneumococcal growth and identify pneumococcal genomic variants associated with growth kinetics for each condition.

Considering that our dataset contained only a third of the >100 known pneumococcal serotypes globally3; due to the geographical variation in the distribution of pneumococcal serotypes80, future studies should investigate all serotypes to provide further insights into the impact of different serotypes on pneumococcal growth kinetics.

Additionally, considering that we generated the phylogeny of the entire pneumococcal species rather than isolates belonging to a single strain or lineage, as defined by several approaches11,81,82, we did not remove recombinations using widely used methods designed for clonal bacterial populations, such as Gubbins83. Therefore, there is a possibility that some branch lengths in the phylogeny used for ancestral trait reconstruction analysis may be less accurate, although the topology is usually robust to recombination84. Furthermore, there may be other variants that potentially influence the pneumococcal growth kinetics, which are regarded as not statistically significant in this study based on our conservative genome-wide significance threshold. As more similar studies are conducted, particularly those utilising much larger datasets, potential variants associated with pneumococcal growth kinetics may be discovered.

Our analysis of population-level genomic and high-throughput in vitro pneumococcal growth data suggests that the capsular serotype and lineage influence pneumococcal growth. However, although some growth features are highly heritable, there appears to be no specific genetic variants that strongly influence these phenotypes, suggesting that a combination of genomic loci with minor effects, individually or collectively, may play a greater role in pneumococcal growth kinetics. Overall, our study provides a proof-of-concept for combining phenotyping and population genomics to understand the genetic basis for pneumococcal growth, an approach which can be applied to study other bacteria of clinical significance and those used in bioengineering, genetics, and the food industry85–87. Ultimately, we anticipate that applying our approach will have implications for devising strategies to reduce life-threatening bacterial infections.

Methods

Ethics statement

The isolates used in this work were collected through an observational cohort study Pneumococcal Bacteraemia Collection Nijmegen (PBCN)37. The non-applicability of the Medical Research Involving Human Subjects Act (WMO) on the study design was confirmed by the Regional Ethics Committee, and the local medical ethics committees of the participating hospitals in the Netherlands approved the study procedures. Study procedures were approved by the Medical Ethical committees of the participating hospitals, including a waiver for individual informed consent (file number: 2020–6644 Radboudumc).

Sample characteristics and whole-genome sequencing

Three hundred forty-eight (n=348) pneumococcal isolates collected from patients with invasive pneumococcal disease through the PBCN observational cohort study37 with in vitro growth data were available for the study. We included a single isolate per patient as invasive pneumococcal infections are typically clonal, so a single isolate is sufficient. The consecutive blood isolates were collected at two hospitals in Nijmegen, the Netherlands, before (2000-2006) and after (2007-2011) the introduction of the PCV7 vaccine into the Dutch paediatric immunisation programme37. Most of the bacteremia cases (97%) concerned adults. Additional information on the PBCN study, including collected clinical variables and phenotypic in vitro growth data, has been previously described17,37,39,40. To generate whole genome sequences, we sequenced a single isolate per patient using an Illumina HiSeq sequencer (Illumina, San Diego, CA, USA). We then assembled the generated reads using SPAdes genome assembler (version 3.14.0)88, and evaluated the draft assemblies using assembly-stats (version: 1.0.1) (https://github.com/sanger-pathogens/assembly-stats). In addition, we used a combination of metrics, including genome size and the number of contigs, to assess the quality of the draft assemblies for inclusion in the analysis.

Measurement of in vitro pneumococcal growth kinetics and derivation of growth features

We measured the pneumococcal growth kinetics for slightly over 15 hours as described by Arends et al.17. In brief, we incubated a frozen aliquot of standardised pneumococcal inoculum in 1.5 ml of blood-like medium (50% M17, 50% casamino acids tryptone [CAT] medium, 0.25% glucose). We measured growth kinetics at 37°C and 5% CO2 every 10 minutes for over 15 hours at an OD620 using a humidity cassette with a microplate reader (Spark 10M, Tecan, Switzerland). We added 15 μL of inoculum was added to 1.5 mL of prewarmed rich growth medium (45% M17, 45% CAT, 0.225% glucose, 10% fetal calf serum [Greiner Bio-one], 26 U/mL catalase [Sigma-Aldrich C1345]) in each well of a sterile flat-bottomed 48-well plate (Nunclon Surface, Nunc, Denmark). The growth medium was supplemented with catalase. We selected these nutrient-rich conditions to highlight the differences in the growth features of the pneumococcus. We performed six repeat measurements on three separate days for each isolate. To understand the growth kinetics of the invasive Dutch pneumococcal isolates, we fitted a modified logistic growth curve function, i.e.+◻0 +◻+◻+◻+◻/(1 ++◻−+◻(+◻–+◻0)), separately to the growth data for each isolate and replicate, using the “optim” function in stats (version 4.3.2) R package (https://www.R-project.org/). We then averaged the inferred values for each replicate to obtain average estimates for each parameter. The parameter L0 represents the initial cell density measured as absorbance or optical density at a wavelength of 620 nm (OD620); Lmaxis the maximum cell density; t0 is the initial time, and r is the average growth rate. We fitted two curves in total for each isolate based on the averaged data from the six replicates using the limited-memory Broyden-Fletcher-Goldfarb-Shanno (L-BFGS-B) optimisation algorithm by minimising the residual sum of squares89. We fitted the first curve to all the data points and the second curve to the data measured until the end of the stationary phase or the beginning of the death phase, whichever came earlier. We defined the end of the stationary phase or the start of the death phase as the time when the optical cell density decreased by more than 5% from the maximum cell density. The curve fitted based on the data measurements collected up to the end of the stationary phase or death phase, or the end of the experiment, if none of these phases had been reached, captured the growth dynamics better than the curve fitted using all the data points to accurately estimate the growth rate and the maximum cell density corresponding to the peak of the pneumococcal growth curve, especially for growth curves with observably short stationary phases and rapid autolysis. Therefore, parameters for the downstream analyses were inferred from the curves fitted to the reduced dataset.

We determined four growth phenotypes for each pneumococcal isolate estimated from the fitted growth curve, namely the lag-phase duration, maximum growth density (Hmax), i.e., captured by parameter Lmax, maximum growth change (ΔH) corresponding to the difference of the maximum and minimum density (Lmax-L0), and average growth rate (r). We defined the lag-phase duration as the time when the optical cell density was at least 1.5-fold higher than the initial cell density at the start of the experiment, as described by Arends et al.17. Due to the challenges of characterising the kinetics at the stationary or death phase, partly due to the high strain-to-strain variability or short study period, we were unable to derive the growth features that capture these phases, including the duration of the stationary and death phases. The logistic growth curves showed an excellent fit for the data, providing an opportunity to investigate the impact of capsular serotype and genetic background (or lineage), as well as specific genomic loci, independent of the genetic background, on the estimated growth parameters and features.

Molecular typing of sequenced isolates

Molecular typing of the isolates to determine capsular serotypes and sequence types (STs) was done in silico using Pathogenwatch (https://pathogen.watch/), a global platform for genomic surveillance90. We determined the capsular serotypes and STs using SeroBA (version v1.0.1)91 and the multilocus sequence typing (MLST) scheme for the pneumococcus38, respectively.

Pneumococcal lineages were defined based on the global pneumococcal sequence cluster (GPSC) nomenclature11 using PopPUNK (version 1.1.0)92. Additionally, we checked the species of each sequenced isolate using Speciator (version 3.0.1) in Pathogenwatch.

Construction of multiple sequence alignments and phylogenetic trees

We mapped the generated sequence reads for each isolate and a draft assembly of Streptococcus oralis strain SF100 genome (GenBank accession: NZ_CP069427) to the pneumococcal ATCC 700669 reference genome (GenBank accession: NC_011900) using Snippy (version 4.6.0) (https://github.com/tseemann/snippy). We identified 155,305 SNPs in the generated alignment using SNP-sites (version 2.3.2)93 and then used them to construct a maximum-likelihood phylogeny using IQ-TREE (version 2.0.3)94. We used the alignment of variable sites by mapping to a reference genome to generate a maximum-likelihood phylogeny using IQ-TREE, as it has been shown to produce more accurate phylogenies95. The general time reversible (GTR) nucleotide substitution model and Gamma heterogeneity of the substitution rates among sites.

We assessed branch support using the ultrafast bootstrap approximation specified by the “-B 1000” option96. We optimised bootstrapped trees using nearest neighbour interchange method on bootstrap alignments by selecting the “--bnni” option. We rooted the phylogenetic tree of the pneumococcal isolates using Streptococcus oralis strain SF100 genome as an outgroup (not shown in the phylogeny). An annotated and interactive version of the phylogeny is available on the Microreact web tool97 (https://microreact.org/project/wxmdblfgprepwugvbbtaq7).

Quantifying the phylogenetic signal

To assess the correlation between the estimated growth features and pneumococcal phylogeny, we used models of continuous character evolution to estimate the phylogenetic signal using Pagel’s λ42 implemented in the “phylosig” function in phytools package (version 2.1.1)98. Pagel’s λ transforms the internal branches relative to the terminal branches and generates values typically ranging from zero to one, with smaller values indicating no phylogenetic signal or independent evolution of the phenotype from the phylogeny. In comparison, higher values close to one show a high phylogenetic signal or a decreased phenotype variability among genetically similar taxa at the tips of the phylogeny. First, we transformed the estimated growth features to a normal distribution using a rank-based inverse normal transform implemented in the RNOmni package (version 1.0.1.2) (https://cran.r-project.org/web/packages/RNOmni/). Next, we mapped the estimated continuous growth feature values onto the phylogeny using fast estimation of maximum likelihood for ancestral states98.

Bacterial GWAS

We identified the presence and absence patterns of unitig sequences to capture genetic variation in the pneumococcal genomes included in this study for the GWAS analysis. We identified unitig sequences from 31bp k-mers using Bifrost (version 1.0.6)99. First, we inferred unitig sequences from a compacted De Bruijn graph constructed using k-mers of length 31, based on draft assemblies of all the pneumococcal isolates included in this study. We generated De Bruijn graphs for each isolate separately and then queried them with the unitig sequences found in the entire dataset to identify the presence and absence of each unitig found in each sequenced genome. We considered a unitig present when 100% of the k-mer was presented in the De Bruijn graph. We combined the presence and absence patterns of the unitig sequences from all the isolates to construct a single matrix.

Since the linear mixed effects models require the tested continuous phenotype to be normally distributed, we used the normalised growth features after applying the rank-based inverse normal transformation using the RNOmni package (version 1.0.1.2) (https://cran.r-project.org/web/packages/RNOmni/). We performed GWAS analysis of pneumococcal growth features using robust linear mixed-effects models implemented in Pyseer (version 1.3.7-dev)100. A kinship matrix specifying the variances and covariances of the random effects for each isolate was generated using the “similarity_pyseer” script in Pyseer. Using this kinship matrix, Pyseer fits an unobserved random effect for each isolate to account for the clonal population structure in the GWAS. We excluded unitigs with <5% minimum allele frequency and >95% maximum allele frequency. The inferred statistical significance for each variant was adjusted to correct for multiple testing using the Bonferroni correction. We estimated the total number of unique unitig presence and absence patterns in the dataset using a script included in Pyseer (https://github.com/mgalardini/pyseer/blob/master/scripts/count_patterns.py). We used the obtained value when adjusting for multiple testing using the Bonferroni correction approach.

Therefore, we defined the genome-wide statistical significance threshold as P <2.58×10−07, i.e., /G, where =0.05 and G=194,053 is the number of unique unitig presence and absence patterns, while the suggestive threshold is P<5.15×10−06, whereby =1. To check potential issues arising from the population structure, we generated quantile-quantile (QQ) plots from the output of GWAS analysis for each growth feature to compare the expected and observed P-values using qqman (version 0.1.7)101. The overall proportion of phenotypic variability explained by variation in the genome (narrow-sense heritability [h2]) was estimated using GEMMA (version 0.98.1)102 and Pyseer. The input files for GEMMA were prepared based on the unitig presence and absence data using PLINK (version 1.90b4) 103. To identify variants associated with the growth features independently of the serotype, we repeated the GWAS by including the serotype as a covariate.

Annotation of variants identified by the GWAS analysis

To annotate the unitig sequences with statistical significance values below the suggestive and genome-wide significance thresholds, we used a custom BioPython 104 wrapper script for BLASTN (version 2.5.0+)105 to identify genomic features in the ATCC 700669 reference genome (GenBank accession: NC_011900) and other complete pneumococcal genomes downloaded from GenBank105. The GWAS results were summarised and visualised using Manhattan plots in R (version 4.0.3) [https://www.R-project.org/].

Other statistical analysis

We compared the growth features between serotypes and lineages using the Kruskal-Wallis non-parametric test. The distribution of the growth features was plotted using violin plots using the vioplot package [version 0.4.0] (https://CRAN.R-project.org/package=vioplot). To assess the contribution of serotype and lineage to the variability in the pneumococcal growth features, we used linear mixed-effects models implemented in the lme4 package (version 1.1.35.1)106 were used, with each growth feature specified as the outcome and the serotype and lineage as random effects. The statistical significance for the linear mixed model was inferred using the lmerTest package [version 3.1.3] (https://CRAN.R-project.org/package=lmerTest). To assess the contribution of serotype and lineage to the variability of each growth feature, we calculated the proportion of variance attributable to the serotype and lineage components of the total variance. We calculated the average invasiveness of the serotypes based on a meta-analysis of data collected in different countries in infants before the introduction of the PCVs50 using the random-effects model implemented in the metafor package [version 2.4.0] (https://CRAN.R-project.org/package=metafor). We defined invasiveness as the odds ratio of detecting a pneumococcal serotype among isolates causing invasive disease compared to nasopharyngeal carriage. We added a value of one to cells with zero counts. Due to the unavailability of the nasopharyngeal carriage data, we imputed the prevalence of serotypes in the carriage by dividing the prevalence in invasive disease by the average invasiveness of each serotype. Next, we used a Spearman correlation test to assess the association between the mean prevalence of serotypes in the nasopharyngeal carriage and invasive disease, and average invasiveness with the average in vitro pneumococcal growth rate.

Data availability

We have deposited the whole-genome sequencing data for the study isolates in the European Nucleotide Archive (ENA) and provided the accession numbers and other isolate metadata in Supplementary Data 1. All the other data supporting the findings of this study are described in this paper or are available as part of the supplementary material. We have also provided additional raw data and code used for the analysis on GitHub: https://github.com/ChrispinChaguza/SpnGrowthKinetics.